---
title: ""
message: false
warning: false
---

# Packages

```{r}

library(broom.mixed)
library(car)
library(clubSandwich)
library(cowplot)
library(data.table)
library(DescTools)
library(dplyr)
library(dunn.test)
library(emmeans)
library(extrafont)
library(future)
library(future.apply)
library(ggplot2)
library(ggpubr)
library(ggthemes)
library(grid)
library(gridExtra)
library(gtable)
library(merTools)
library(nlme)
library(openxlsx)
library(patchwork)
library(performance)
library(purrr)
library(RColorBrewer)
library(readxl)
library(reshape2)
library(rstatix)
library(sf)
library(scales)
library(spaMM)
library(splines)
library(stringr)
library(tidyr)
library(tidyverse)
library(viridis)
library(writexl)

font_import()           
loadfonts(device = "pdf")

```

# Reading and Preprocessing

```{r,eval = FALSE, echo = TRUE}

all_PF_SF <- fread("/media/amazonia/hd2/lais/GitHub/paper_climate_forest_regeneration/plans_results/plan_final/all_vars_2001-2021_with_age.csv")

n_obs_cba_a <- all_PF_SF %>% 
  filter(Region == "CBA")

n_obs_eba_a <- all_PF_SF %>% 
  filter(Region == "EBA")

n_obs_sba_a <- all_PF_SF %>% 
  filter(Region == "SBA")

gc()

```

```{r,eval = FALSE, echo = TRUE}

all_lulc <- dcast(
  all_PF_SF, 
  Lon + Lat + Prec + Evapo + Temp + Region + Ano + Idade_SF ~ Tipo, 
  value.var = "frac_area", 
  fun.aggregate = sum, 
  fill = 0
)

setnames(all_lulc, old = c("PF", "SF", "Agriculture", "Pasture"), 
                  new = c("frac_area_PF", "frac_area_SF", "frac_area_Agriculture", "frac_area_Pasture"))

rm(all_PF_SF)

```

## Filtering – "Non-pure" Pixels

```{r,eval = FALSE, echo = TRUE}

all_lulc_fil_50 <- all_lulc %>%
  filter(
    (frac_area_SF > 0.5 & frac_area_PF == 0) |

    (frac_area_SF > 0 & frac_area_SF <= 0.5 & 
     (frac_area_PF + frac_area_Agriculture + frac_area_Pasture) < frac_area_SF) |
    
    (frac_area_SF == 0 & frac_area_PF > 0.5) |
    
    (frac_area_PF > 0 & frac_area_PF <= 0.5 & 
     (frac_area_Agriculture + frac_area_Pasture + frac_area_SF) < frac_area_PF) |

    (frac_area_SF == 0 & frac_area_PF == 0 & frac_area_Agriculture >= 0.1) |

    (frac_area_SF == 0 & frac_area_PF == 0 & frac_area_Pasture >= 0.1)
    )

fwrite(all_lulc_fil_50, file = "/media/amazonia/hd2/lais/GitHub/paper_climate_forest_regeneration/data_group/plan_fil_all_lulc.csv")

```

```{r,eval = FALSE, echo = TRUE}

all_lulc_long_50 <- melt(all_lulc_fil_50, 
                         id.vars = c("Lon", "Lat", "Prec", "Evapo", "Temp", "Region", "Ano", "Idade_SF"),
                         variable.name = "Tipo", 
                         value.name = "frac_area")

setDT(all_lulc_long_50)

all_lulc_long_50[, Tipo := gsub("frac_area_", "", Tipo)]

all_lulc_50_final <- all_lulc_long_50[frac_area > 0.01]

fwrite(all_lulc_50_final, file = "/media/amazonia/hd2/lais/GitHub/paper_climate_forest_regeneration/data_group/plan_all_lulc_final.csv")

rm(all_lulc)
rm(all_lulc_fil_50)
rm(all_lulc_long_50)

gc()

```

```{r,eval = FALSE, echo = TRUE}

n_obs_cba_d <- all_lulc_50_final %>% 
  filter(Region == "CBA")

n_obs_eba_d <- all_lulc_50_final %>% 
  filter(Region == "EBA")

n_obs_sba_d <- all_lulc_50_final %>% 
  filter(Region == "SBA")

```

## Plot - Observations

```{r,eval = FALSE, echo = TRUE}

before <- c(
  CBA = nrow(n_obs_cba_a),
  EBA = nrow(n_obs_eba_a),
  SBA = nrow(n_obs_sba_a)
)

after <- c(
  CBA = nrow(n_obs_cba_d),
  EBA = nrow(n_obs_eba_d),
  SBA = nrow(n_obs_sba_d)
)

reducoes_percentuais <- round((before - after) / before * 100, 1)
regioes <- c("CBA", "EBA", "SBA")

df_plot <- data.frame(
  Regiao = rep(regioes, each = 2),
  Status = factor(rep(c("Before", "After"), times = 3), levels = c("Before", "After")),
  Observacoes = as.vector(rbind(before, after))
)

df_text <- df_plot %>%
  group_by(Regiao) %>%
  summarize(y_pos = max(Observacoes) * 1.05) %>%
  mutate(
    x_pos = Regiao,
    Reducao = paste0("-", reducoes_percentuais, "%")
  )

ggplot(df_plot, aes(x = Regiao, y = Observacoes, fill = Status)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.7), width = 0.6) +
  geom_text(data = df_text, aes(x = x_pos, y = y_pos, label = Reducao),
            inherit.aes = FALSE, size = 3, fontface = "bold", family = "serif") +
  scale_fill_manual(values = c("Before" = "skyblue", "After" = "lightgreen")) +
  labs(
    title = "",
    y = "Number of observations (pixels)", x = "", fill = ""
  ) +
  ggthemes::theme_few()+
  theme(legend.position = c(0.15, 0.90),
    legend.background = element_rect(fill = "transparent"),
    text = element_text(size = 12, face = "bold", family = "serif"),
    axis.text = element_text(size = 10, face = "plain", family = "serif", color = "black"),
    axis.title.y = element_text(size = 10))

#ggsave("fig/observatios_before_after.png", dpi = 600)

```

```{r,eval = FALSE, echo = TRUE}

rm(n_obs_cba_a, n_obs_eba_a, n_obs_sba_a)
rm(n_obs_cba_d, n_obs_eba_d, n_obs_sba_d)
rm(df_plot, df_text)

gc()

```

# Descriptive Analyses

## PRECIPITATION

### Data reading

```{r,eval = FALSE, echo = TRUE}

all_lulc_50_final <- fread("/media/amazonia/hd2/lais/GitHub/paper_climate_forest_regeneration/data_group/plan_all_lulc_final.csv")

all_lulc_50_fil <- all_lulc_50_final %>% 
   filter(Tipo %in% c("SF", "PF"))

rm(all_lulc_50_final)

n_obs_cba_d <- all_lulc_50_fil %>% 
  filter(Region == "CBA")

n_obs_eba_d <- all_lulc_50_fil %>% 
  filter(Region == "EBA")

n_obs_sba_d <- all_lulc_50_fil %>% 
  filter(Region == "SBA")

gc()

```

### Grouping

```{r,eval = FALSE, echo = TRUE}

all_lulc_prec <- all_lulc_50_fil %>% 
  transmute(
    Prec = Prec,
    Frac = frac_area,
    Tipo = Tipo,
    Ano = Ano,
    region = Region,
    Idade = Idade_SF
  ) %>%
  mutate(
    ff_area = case_when(Frac >= 0.01 & Frac <= 0.1 ~ "1-10",
                        Frac > 0.1 & Frac <= 0.2 ~ "10-20",
                        Frac > 0.2 & Frac <= 0.3 ~ "20-30",
                        Frac > 0.3 & Frac <= 0.4 ~ "30-40",
                        Frac > 0.4 & Frac <= 0.5 ~ "40-50",
                        Frac > 0.5 & Frac <= 0.6 ~ "50-60",
                        Frac > 0.6 & Frac <= 0.7 ~ "60-70",
                        Frac > 0.7 & Frac <= 0.8 ~ "70-80",
                        Frac > 0.8 & Frac <= 0.9 ~ "80-90",
                        Frac > 0.9 & Frac <= 1.0 ~ "90-100")) %>%
  mutate(
    idade = case_when(Idade >= 0 & Idade <= 10 ~ "<= 10",
                      Idade > 10 & Idade <= 20 ~ "10-20",
                      Idade > 20 ~ "> 20")) %>%
  group_by(Ano, Tipo, region, ff_area, idade) %>% 
  summarise(
    Prec = mean(Prec),
    n_obs = n(),
    .groups = "drop"
  )


all_lulc_prec_stages <- all_lulc_prec %>%
  mutate(Tipo = case_when(Tipo == "PF" ~ "PF",
                          Tipo == "SF" & idade == "<= 10" ~ "SF_early",
                          Tipo == "SF" & idade == "10-20" ~ "SF_intermediate",
                          Tipo == "SF" & idade == "> 20" ~ "SF_advanced",
                          TRUE ~ NA_character_)) %>% 
  filter(!is.na(Tipo))  


all_lulc_prec_final <- all_lulc_prec_stages %>% 
  mutate(type = case_when(Tipo == "PF" ~ 1,
                          Tipo == "SF_early" ~ 2, 
                          Tipo == "SF_intermediate" ~ 3,
                          Tipo == "SF_advanced" ~ 4))


max_min_prec = all_lulc_prec_final %>%
  group_by(Tipo, region, ff_area, idade) %>%
  summarise(
    max_prec = max(Prec),
    median_prec = median(Prec),
    min_prec = min(Prec)
  )

write_xlsx(max_min_prec, "data_group/plan_max_and_min_group_prec.xlsx")

write_xlsx(all_lulc_prec_final, "data_group/plan_data_group_prec.xlsx")

rm(all_lulc_prec, all_lulc_prec_stages)
rm(max_min_prec)

gc()

```

### Tests for Normality, Homoscedasticity, and Independence

```{r}

all_pf_sf_prec = read_xlsx("data_group/plan_data_group_prec.xlsx")

shapiro_test_results_prec <- all_pf_sf_prec %>%
  group_by(Tipo, region, ff_area) %>%
  summarise(
    p_value = shapiro.test(Prec)$p.value
  ) %>%
  ungroup()

print(shapiro_test_results_prec)

write_xlsx(shapiro_test_results_prec, "data_group/shapiro_test_prec_results.xlsx")


all_pf_sf_prec$Tipo <- as.factor(all_pf_sf_prec$Tipo)

results_list_prec <- list()

for (region in unique(all_pf_sf_prec$region)) {
  for (ff_area in unique(all_pf_sf_prec$ff_area)) {
    
    data_subset <- all_pf_sf_prec %>%
      filter(region == !!region, ff_area == !!ff_area)
    
    if (nrow(data_subset) > 1 && length(unique(data_subset$Tipo)) > 1) {
      try({
        levene_test_prec <- leveneTest(Prec ~ Tipo, data = data_subset)
        p_value <- levene_test_prec$`Pr(>F)`[1]  
      }, silent = TRUE)
    } else {
      p_value <- NA  
    }
    
    if (!is.null(p_value)) {
      results_list_prec[[length(results_list_prec) + 1]] <- data.frame(region = region, 
                                                             ff_area = ff_area, 
                                                             p_value = p_value)
    }
  }
}

levene_test_results_prec <- do.call(rbind, results_list_prec)

print(levene_test_results_prec)

write_xlsx(levene_test_results_prec, "data_group/levene_test_prec_results.xlsx")


kendall_results_list_prec <- list()

for (region in unique(all_pf_sf_prec$region)) {
  for (ff_area in unique(all_pf_sf_prec$ff_area)) {
    
    data_subset <- all_pf_sf_prec %>%
      filter(region == !!region, ff_area == !!ff_area)

    if (nrow(data_subset) > 1 && length(unique(data_subset$Tipo)) > 1) {
      try({
        cor_test <- cor.test(as.numeric(data_subset$Prec), as.numeric(data_subset$Tipo), method = "kendall")
        p_value <- cor_test$p.value  
        cor_value <- cor_test$estimate  
      }, silent = TRUE)
    } else {
      p_value <- NA  
      cor_value <- NA
    }
    
    if (!is.null(p_value)) {
      kendall_results_list_prec[[length(kendall_results_list_prec) + 1]] <- data.frame(
        region = region, 
        ff_area = ff_area, 
        cor_value = cor_value,
        p_value = p_value
      )
    }
  }
}

kendall_test_results_prec <- do.call(rbind, kendall_results_list_prec)

write_xlsx(kendall_test_results_prec, "data_group/kendall_test_prec_results.xlsx")

gc()

```

### t-test between coverage types – PF and SF

```{r}

area_ranges <- c("1-10", "10-20", "20-30", "30-40", "40-50", 
                 "50-60", "60-70", "70-80", "80-90", "90-100")

regions <- c("CBA", "EBA", "SBA")

significance_results_t_prec <- data.frame(Intervalo = area_ranges)

for (region in regions) {
  
  p_values_pf_sf_early <- c()
  p_values_pf_sf_intermediate <- c()
  p_values_pf_sf_advanced <- c()

  for (range in area_ranges) {
    
    pv_data <- all_pf_sf_prec %>% filter(region == !!region, ff_area == range, Tipo == "PF")
    sv_early_data <- all_pf_sf_prec %>% filter(region == !!region, ff_area == range, Tipo == "SF_early")
    sv_intermediate_data <- all_pf_sf_prec %>% filter(region == !!region, ff_area == range, Tipo == "SF_intermediate")
    sv_advanced_data <- all_pf_sf_prec %>% filter(region == !!region, ff_area == range, Tipo == "SF_advanced")

    if (nrow(pv_data) > 1 && nrow(sv_early_data) > 1) {
      p_values_pf_sf_early <- c(p_values_pf_sf_early, t.test(pv_data$Prec, sv_early_data$Prec, var.equal = TRUE)$p.value)
    } else {
      p_values_pf_sf_early <- c(p_values_pf_sf_early, NA)
    }
    
    if (nrow(pv_data) > 1 && nrow(sv_intermediate_data) > 1) {
      p_values_pf_sf_intermediate <- c(p_values_pf_sf_intermediate, t.test(pv_data$Prec, sv_intermediate_data$Prec, var.equal = TRUE)$p.value)
    } else {
      p_values_pf_sf_intermediate <- c(p_values_pf_sf_intermediate, NA)
    }
    
    if (nrow(pv_data) > 1 && nrow(sv_advanced_data) > 1) {
      p_values_pf_sf_advanced <- c(p_values_pf_sf_advanced, t.test(pv_data$Prec, sv_advanced_data$Prec, var.equal = TRUE)$p.value)
    } else {
      p_values_pf_sf_advanced <- c(p_values_pf_sf_advanced, NA)
    }
  }
  
  significance_results_t_prec[[paste0(region, "_PF_vs_SF_early")]] <- p_values_pf_sf_early
  significance_results_t_prec[[paste0(region, "_PF_vs_SF_intermediate")]] <- p_values_pf_sf_intermediate
  significance_results_t_prec[[paste0(region, "_PF_vs_SF_advanced")]] <- p_values_pf_sf_advanced
}

write_xlsx(significance_results_t_prec, "data_group/significance_results_prec_testt.xlsx")

gc()

```

### Wilcoxon test between coverage types – PF and SF

```{r}

all_pf_sf_prec = read_xlsx("data_group/plan_data_group_prec.xlsx")

area_ranges <- c("1-10", "10-20", "20-30", "30-40", "40-50",
                 "50-60", "60-70", "70-80", "80-90", "90-100")

regions <- c("CBA", "EBA", "SBA")

significance_results_w_prec <- data.frame(Intervalo = area_ranges)

for (region in regions) {
  
  p_values_pf_sf_early <- c()
  p_values_pf_sf_intermediate <- c()
  p_values_pf_sf_advanced <- c()

  for (range in area_ranges) {

    pv_data <- all_pf_sf_prec %>% filter(region == !!region, ff_area == range, Tipo == "PF")
    sv_early_data <- all_pf_sf_prec %>% filter(region == !!region, ff_area == range, Tipo == "SF_early")
    sv_intermediate_data <- all_pf_sf_prec %>% filter(region == !!region, ff_area == range, Tipo == "SF_intermediate")
    sv_advanced_data <- all_pf_sf_prec %>% filter(region == !!region, ff_area == range, Tipo == "SF_advanced")

    if (nrow(pv_data) > 1 && nrow(sv_early_data) > 1) {
      p_values_pf_sf_early <- c(p_values_pf_sf_early, wilcox.test(pv_data$Prec, sv_early_data$Prec, paired = FALSE)$p.value)
    } else {
      p_values_pf_sf_early <- c(p_values_pf_sf_early, NA)
    }
    
    if (nrow(pv_data) > 1 && nrow(sv_intermediate_data) > 1) {
      p_values_pf_sf_intermediate <- c(p_values_pf_sf_intermediate, wilcox.test(pv_data$Prec, sv_intermediate_data$Prec, paired = FALSE)$p.value)
    } else {
      p_values_pf_sf_intermediate <- c(p_values_pf_sf_intermediate, NA)
    }
    
    if (nrow(pv_data) > 1 && nrow(sv_advanced_data) > 1) {
      p_values_pf_sf_advanced <- c(p_values_pf_sf_advanced, wilcox.test(pv_data$Prec, sv_advanced_data$Prec, paired = FALSE)$p.value)
    } else {
      p_values_pf_sf_advanced <- c(p_values_pf_sf_advanced, NA)
    }
  }

  significance_results_w_prec[[paste0(region, "_PF_vs_SF_early")]] <- p_values_pf_sf_early
  significance_results_w_prec[[paste0(region, "_PF_vs_SF_intermediate")]] <- p_values_pf_sf_intermediate
  significance_results_w_prec[[paste0(region, "_PF_vs_SF_advanced")]] <- p_values_pf_sf_advanced
}

write_xlsx(significance_results_w_prec, "data_group/significance_results_prec_Wilcox_test.xlsx")

gc()

```

#### Plot with significance

```{r}

long_significance_results_w_prec <- significance_results_w_prec %>%
  gather(key = "Comparison", value = "P_value", -Intervalo) %>%
  separate(Comparison, into = c("region", "Comparison_Type"), sep = "_PF_vs_")

all_pf_sf_prec <- all_pf_sf_prec %>%
  left_join(long_significance_results_w_prec, 
            by = c("ff_area" = "Intervalo", "region" = "region", "Tipo" = "Comparison_Type"))

all_pf_sf_prec$Significant <- ifelse(all_pf_sf_prec$P_value < 0.05, "*", "")

all_pf_sf_prec <- all_pf_sf_prec %>%
  mutate(Tipo = factor(Tipo, levels = c("PF", "SF_early", "SF_intermediate", "SF_advanced")))

get_whisker_max <- function(x) {
  stats <- boxplot.stats(x)$stats
  return(stats[5])
}

significance_positions <- all_pf_sf_prec %>%
  group_by(ff_area, region, Tipo) %>%
  summarise(max_whisker = get_whisker_max(Prec), .groups = "drop") %>%
  left_join(long_significance_results_w_prec, 
            by = c("ff_area" = "Intervalo", "region" = "region", "Tipo" = "Comparison_Type")) %>%
  filter(P_value < 0.05) %>%
  mutate(
    asterisk_y = max_whisker + 15, 
    x_adjust = case_when(
      Tipo == "SF_early" ~ -0.1,
      Tipo == "SF_intermediate" ~ 0.1,
      Tipo == "SF_advanced" ~ 0.3
    ),
    x_pos = as.numeric(factor(ff_area, levels = area_ranges)) + x_adjust
  )

p_ast_prec_w <- all_pf_sf_prec %>%
  ggplot(aes(x = ff_area, y = Prec, fill = Tipo)) +
  geom_boxplot(outlier.alpha = 0.4) +
  facet_grid(~region) + 
  ggthemes::theme_few() +
  theme(
    legend.position = c(0.77, 0.84),
    legend.background = element_rect(fill = "transparent"),
    text = element_text(size = 12, face = "bold", family = "serif"), 
    axis.text = element_text(size = 10, face = "plain", family = "serif", color = "black"),
    axis.text.x = element_blank(),
    legend.text = element_text(size = 9),
    panel.spacing.x = unit(1, "lines"), 
    panel.spacing.y = unit(1, "lines"),
    axis.title = element_text(size = 14, margin = margin(r = 20)),
    strip.text = element_text(size = 14, face = "bold", family = "serif")
  ) +
  scale_fill_manual(values = c("midnightblue", "#87CEFA", "#1E90FF", "#0000FF"), name = NULL, labels = c("PF", "SF early", "SF intermediate", "SF advanced")) +
  scale_y_continuous(breaks = seq(0, max(all_pf_sf_prec$Prec, na.rm = TRUE), by = 250)) +
  labs(
    x = "",
    y = "Average annual accumulated \n precipitation (mm)",
    #y = expression(atop("Average annual accumulated", paste("precipitation (mm·", yr^{-1}, ")")))
  ) +
  geom_text(data = significance_positions, aes(x = x_pos, y = asterisk_y, label = "*"),
            size = 3, color = "red")

p_ast_prec_w

ggsave("fig/area_x_prec_pf_sf_significant_asterisco_wilcoxon.png", dpi = 600, width = 9, height = 5.5)

gc()

```

## EVAPOTRANSPIRATION

### Grouping

```{r,eval = FALSE, echo = TRUE}

all_lulc_50_final = fread("/media/amazonia/hd2/lais/GitHub/paper_climate_forest_regeneration/data_group/plan_all_lulc_final.csv")

all_lulc_50_fil <- all_lulc_50_final %>%
   filter(Tipo %in% c("SF", "PF"))

all_lulc_evapo = all_lulc_50_fil %>%
  filter(!Evapo >= 2000) %>% 
  transmute(
    Evapo = Evapo,
    Frac = frac_area,
    Tipo = Tipo,
    Ano = Ano,
    region = Region,
    Idade = Idade_SF
  ) %>%
  mutate(
    ff_area = case_when(Frac >= 0.01 & Frac <= 0.1 ~ "1-10",
                        Frac > 0.1 & Frac <= 0.2 ~ "10-20",
                        Frac > 0.2 & Frac <= 0.3 ~ "20-30",
                        Frac > 0.3 & Frac <= 0.4 ~ "30-40",
                        Frac > 0.4 & Frac <= 0.5 ~ "40-50",
                        Frac > 0.5 & Frac <= 0.6 ~ "50-60",
                        Frac > 0.6 & Frac <= 0.7 ~ "60-70",
                        Frac > 0.7 & Frac <= 0.8 ~ "70-80",
                        Frac > 0.8 & Frac <= 0.9 ~ "80-90",
                        Frac > 0.9 & Frac <= 1.0 ~ "90-100")) %>% 
  mutate(
    idade = case_when(Idade >= 0 & Idade <= 10 ~ "<= 10",
                      Idade > 10 & Idade <= 20 ~ "10-20",
                      Idade > 20 ~ "> 20")) %>%
  group_by(Ano, Tipo, region, ff_area, idade) %>% 
  summarise(
    Evapo = mean(Evapo),
    n_obs = n(),
    .groups = "drop"
  )

all_lulc_evapo_stages <- all_lulc_evapo %>%
  mutate(Tipo = case_when(Tipo == "PF" ~ "PF", 
                          Tipo == "SF" & idade == "<= 10" ~ "SF_early", 
                          Tipo == "SF" & idade == "10-20" ~ "SF_intermediate", 
                          Tipo == "SF" & idade == "> 20" ~ "SF_advanced",
                          TRUE ~ NA_character_ )) %>% 
  filter(!is.na(Tipo))  


all_lulc_evapo_final <- all_lulc_evapo_stages %>% 
  mutate(type = case_when(Tipo == "PF" ~ 1,
                          Tipo == "SF_early" ~ 2, 
                          Tipo == "SF_intermediate" ~ 3,
                          Tipo == "SF_advanced" ~ 4))


max_min_evapo = all_lulc_evapo_final %>%
  group_by(Tipo, region, ff_area, idade) %>%
  summarise(
    max_evapo = max(Evapo),
    median_evapo = median(Evapo),
    min_evapo = min(Evapo)
  )

write_xlsx(max_min_evapo, "data_group/plan_max_and_min_group_evapo.xlsx")

write_xlsx(all_lulc_evapo_final, "data_group/plan_data_group_evapo.xlsx")

rm(all_lulc_50_final)

gc()

```

### Tests for Normality, Homoscedasticity, and Independence

```{r}

all_pf_sf_evapo = read_xlsx("data_group/plan_data_group_evapo.xlsx")

shapiro_test_results_evapo <- all_pf_sf_evapo %>%
  group_by(Tipo,region, ff_area) %>%
  summarise(
    p_value = shapiro.test(Evapo)$p.value
  ) %>%
  ungroup()

print(shapiro_test_results_evapo)

write_xlsx(shapiro_test_results_evapo, "data_group/shapiro_test_evapo_results.xlsx")


all_pf_sf_evapo$Tipo <- as.factor(all_pf_sf_evapo$Tipo)

results_list_evapo <- list()

for (region in unique(all_pf_sf_evapo$region)) {
  for (ff_area in unique(all_pf_sf_evapo$ff_area)) {
    
    data_subset <- all_pf_sf_evapo %>%
      filter(region == !!region, ff_area == !!ff_area)
     
    if (nrow(data_subset) > 1 && length(unique(data_subset$Tipo)) > 1) {
      try({
        levene_test_evapo <- leveneTest(Evapo ~ Tipo, data = data_subset)
        p_value <- levene_test_evapo$`Pr(>F)`[1]  
      }, silent = TRUE)
    } else {
      p_value <- NA  
    }
    
    if (!is.null(p_value)) {
      results_list_evapo[[length(results_list_evapo) + 1]] <- data.frame(region = region, 
                                                             ff_area = ff_area, 
                                                             p_value = p_value)
    }
  }
}

levene_test_results_evapo <- do.call(rbind, results_list_evapo)

print(levene_test_results_evapo)

write_xlsx(levene_test_results_evapo, "data_group/levene_test_evapo_results.xlsx")


kendall_results_list_evapo <- list()

for (region in unique(all_pf_sf_evapo$region)) {
  for (ff_area in unique(all_pf_sf_evapo$ff_area)) {
    
    data_subset <- all_pf_sf_evapo %>%
      filter(region == !!region, ff_area == !!ff_area)
    
    if (nrow(data_subset) > 1 && length(unique(data_subset$Tipo)) > 1) {
      try({
        cor_test <- cor.test(as.numeric(data_subset$Evapo), as.numeric(data_subset$Tipo), method = "kendall")
        p_value <- cor_test$p.value  
        cor_value <- cor_test$estimate  
      }, silent = TRUE)
    } else {
      p_value <- NA  
      cor_value <- NA
    }
    
    if (!is.null(p_value)) {
      kendall_results_list_evapo[[length(kendall_results_list_evapo) + 1]] <- data.frame(
        region = region, 
        ff_area = ff_area, 
        cor_value = cor_value,
        p_value = p_value
      )
    }
  }
}

kendall_test_results_evapo <- do.call(rbind, kendall_results_list_evapo)

write_xlsx(kendall_test_results_evapo, "data_group/kendall_test_evapo_results.xlsx")

gc()

```

###t-test between coverage types – PF and SF

```{r}

area_ranges <- c("1-10", "10-20", "20-30", "30-40", "40-50", 
                 "50-60", "60-70", "70-80", "80-90", "90-100")

regions <- c("CBA", "EBA", "SBA")

significance_results_t_evapo <- data.frame(Intervalo = area_ranges)  

for (region in regions) {
  
  p_values_pf_sf_early <- c()
  p_values_pf_sf_intermediate <- c()
  p_values_pf_sf_advanced <- c()
  
  for (range in area_ranges) {
    
    pv_data <- all_pf_sf_evapo %>% filter(region == !!region, ff_area == range, Tipo == "PF")
    sv_early_data <- all_pf_sf_evapo %>% filter(region == !!region, ff_area == range, Tipo == "SF_early")
    sv_intermediate_data <- all_pf_sf_evapo %>% filter(region == !!region, ff_area == range, Tipo == "SF_intermediate")
    sv_advanced_data <- all_pf_sf_evapo %>% filter(region == !!region, ff_area == range, Tipo == "SF_advanced")
    
    if (nrow(pv_data) > 1 && nrow(sv_early_data) > 1) {
      p_values_pf_sf_early <- c(p_values_pf_sf_early, t.test(pv_data$Evapo, sv_early_data$Evapo, var.equal = TRUE)$p.value)
    } else {
      p_values_pf_sf_early <- c(p_values_pf_sf_early, NA)
    }
    
    if (nrow(pv_data) > 1 && nrow(sv_intermediate_data) > 1) {
      p_values_pf_sf_intermediate <- c(p_values_pf_sf_intermediate, t.test(pv_data$Evapo, sv_intermediate_data$Evapo, var.equal = TRUE)$p.value)
    } else {
      p_values_pf_sf_intermediate <- c(p_values_pf_sf_intermediate, NA)
    }
    
    if (nrow(pv_data) > 1 && nrow(sv_advanced_data) > 1) {
      p_values_pf_sf_advanced <- c(p_values_pf_sf_advanced, t.test(pv_data$Evapo, sv_advanced_data$Evapo, var.equal = TRUE)$p.value)
    } else {
      p_values_pf_sf_advanced <- c(p_values_pf_sf_advanced, NA)
    }
  }
  
  significance_results_t_evapo[[paste0(region, "_PF_vs_SF_early")]] <- p_values_pf_sf_early
  significance_results_t_evapo[[paste0(region, "_PF_vs_SF_intermediate")]] <- p_values_pf_sf_intermediate
  significance_results_t_evapo[[paste0(region, "_PF_vs_SF_advanced")]] <- p_values_pf_sf_advanced
}

write_xlsx(significance_results_t_evapo, "data_group/significance_results_evapo_testt.xlsx")

gc()

```

### Wilcoxon test between coverage types – PF and SF

```{r}

all_pf_sf_evapo = read_xlsx("data_group/plan_data_group_evapo.xlsx")

area_ranges <- c("1-10", "10-20", "20-30", "30-40", "40-50", 
                 "50-60", "60-70", "70-80", "80-90", "90-100")

regions <- c("CBA", "EBA", "SBA")

significance_results_w_evapo <- data.frame(Intervalo = area_ranges)

for (region in regions) {
  
  p_values_pf_sf_early <- c()
  p_values_pf_sf_intermediate <- c()
  p_values_pf_sf_advanced <- c()
  
  for (range in area_ranges) {
    
    pv_data <- all_pf_sf_evapo %>% filter(region == !!region, ff_area == range, Tipo == "PF")
    sv_early_data <- all_pf_sf_evapo %>% filter(region == !!region, ff_area == range, Tipo == "SF_early")
    sv_intermediate_data <- all_pf_sf_evapo %>% filter(region == !!region, ff_area == range, Tipo == "SF_intermediate")
    sv_advanced_data <- all_pf_sf_evapo %>% filter(region == !!region, ff_area == range, Tipo == "SF_advanced")
    
    if (nrow(pv_data) > 1 && nrow(sv_early_data) > 1) {
      p_values_pf_sf_early <- c(p_values_pf_sf_early, wilcox.test(pv_data$Evapo, sv_early_data$Evapo, paired = FALSE)$p.value)
    } else {
      p_values_pf_sf_early <- c(p_values_pf_sf_early, NA)
    }
    
    if (nrow(pv_data) > 1 && nrow(sv_intermediate_data) > 1) {
      p_values_pf_sf_intermediate <- c(p_values_pf_sf_intermediate, wilcox.test(pv_data$Evapo, sv_intermediate_data$Evapo, paired = FALSE)$p.value)
    } else {
      p_values_pf_sf_intermediate <- c(p_values_pf_sf_intermediate, NA)
    }
    
    if (nrow(pv_data) > 1 && nrow(sv_advanced_data) > 1) {
      p_values_pf_sf_advanced <- c(p_values_pf_sf_advanced, wilcox.test(pv_data$Evapo, sv_advanced_data$Evapo, paired = FALSE)$p.value)
    } else {
      p_values_pf_sf_advanced <- c(p_values_pf_sf_advanced, NA)
    }
  }
  
  significance_results_w_evapo[[paste0(region, "_PF_vs_SF_early")]] <- p_values_pf_sf_early
  significance_results_w_evapo[[paste0(region, "_PF_vs_SF_intermediate")]] <- p_values_pf_sf_intermediate
  significance_results_w_evapo[[paste0(region, "_PF_vs_SF_advanced")]] <- p_values_pf_sf_advanced
}

write_xlsx(significance_results_w_evapo, "data_group/significance_results_evapo_Wilcox_test.xlsx")

gc()
```

#### Plot with significance

```{r}

long_significance_results_w_evapo <- significance_results_w_evapo %>%
  gather(key = "Comparison", value = "P_value", -Intervalo) %>%
  separate(Comparison, into = c("region", "Comparison_Type"), sep = "_PF_vs_")

all_pf_sf_evapo <- all_pf_sf_evapo %>%
  left_join(long_significance_results_w_evapo, 
            by = c("ff_area" = "Intervalo", "region" = "region", "Tipo" = "Comparison_Type"))

all_pf_sf_evapo$Significant <- ifelse(all_pf_sf_evapo$P_value < 0.05, "*", "")

all_pf_sf_evapo <- all_pf_sf_evapo %>%
  mutate(Tipo = factor(Tipo, levels = c("PF", "SF_early", "SF_intermediate", "SF_advanced")))

get_whisker_max <- function(x) {
  stats <- boxplot.stats(x)$stats
  return(stats[5]) 
}

significance_positions <- all_pf_sf_evapo %>%
  group_by(ff_area, region, Tipo) %>%
  summarise(max_whisker = get_whisker_max(Evapo), .groups = "drop") %>%
  left_join(long_significance_results_w_evapo, 
            by = c("ff_area" = "Intervalo", "region" = "region", "Tipo" = "Comparison_Type")) %>%
  filter(P_value < 0.05) %>%
  mutate(
    asterisk_y = max_whisker + 10, 
    x_adjust = case_when(
      Tipo == "SF_early" ~ -0.1,
      Tipo == "SF_intermediate" ~ 0.1,
      Tipo == "SF_advanced" ~ 0.3
    ),
    x_pos = as.numeric(factor(ff_area, levels = area_ranges)) + x_adjust
  )

p_ast_evapo_w <- all_pf_sf_evapo %>% 
  ggplot(aes(x=ff_area, y=Evapo, fill=Tipo)) +
  geom_boxplot(outlier.alpha = 0.4) +
  facet_grid(~region) + 
  ggthemes::theme_few() +
  theme(
    legend.position = c(0.77, 0.84),
    legend.background = element_rect(fill="transparent"),
    #legend.position = "top",
    text = element_text(size = 12, face = "bold",family = "serif"), 
    #text = element_text(size = 7, face = "bold",family = "serif"), 
    axis.text = element_text(size = 10, face = "plain", family = "serif", color = "black"),
    #axis.text = element_text(size = 6, face = "plain", family = "serif", color = "black"),
    axis.text.x = element_blank(),
    #axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
    legend.text = element_text(size = 9),
    #legend.text = element_text(size = 7),
    panel.spacing.x = unit(1, "lines"),  
    panel.spacing.y = unit(1, "lines"),
    axis.title = element_text(size = 14, margin = margin(r = 20)),
    strip.text = element_blank()) +
  scale_fill_manual(values = c("#02401B", "lightgreen", "green1", "green4"), name = NULL, labels = c("PF", "SF early", "SF intermediate", "SF advanced")) +
  #geom_vline(xintercept = c(1,2,3,4,5,6,7,8,9,10), linetype = "dashed", color = "lightgray", size = 0.5) +
  #labels = c("Primary Vegetation           ", "Secondary Vegetation"))+
  labs(x = "",
       y = "Average annual accumulated \n evapotranspiration (mm)",
       #y = expression(atop("Average annual accumulated", paste("evapotranspiration (mm·", yr^{-1}, ")")))
       ) +
  geom_text(data = significance_positions, 
            aes(x = x_pos, y = asterisk_y, label = "*"),
            size = 3, color = "red")

p_ast_evapo_w

ggsave("fig/area_x_evapo_pf_sf_significant_asterisco_wilcoxon.png", dpi = 600, width = 9, height = 5.5)

gc()

```

## TEMPERTAURE

### Grouping

```{r,eval = FALSE, echo = TRUE}

all_lulc_50_final = fread("/media/amazonia/hd2/lais/GitHub/paper_climate_forest_regeneration/data_group/plan_all_lulc_final.csv")

all_lulc_50_fil <- all_lulc_50_final %>%
   filter(Tipo %in% c("SF", "PF"))

all_lulc_temp = all_lulc_50_fil %>% 
  transmute(
    Temp = Temp,
    Frac = frac_area,
    Tipo = Tipo,
    Ano = Ano,
    region = Region,
    Idade = Idade_SF
  ) %>%
  mutate(
    ff_area = case_when(Frac >= 0.01 & Frac <= 0.1 ~ "1-10",
                        Frac > 0.1 & Frac <= 0.2 ~ "10-20",
                        Frac > 0.2 & Frac <= 0.3 ~ "20-30",
                        Frac > 0.3 & Frac <= 0.4 ~ "30-40",
                        Frac > 0.4 & Frac <= 0.5 ~ "40-50",
                        Frac > 0.5 & Frac <= 0.6 ~ "50-60",
                        Frac > 0.6 & Frac <= 0.7 ~ "60-70",
                        Frac > 0.7 & Frac <= 0.8 ~ "70-80",
                        Frac > 0.8 & Frac <= 0.9 ~ "80-90",
                        Frac > 0.9 & Frac <= 1.0 ~ "90-100")) %>% 
  mutate(
    idade = case_when(Idade >= 0 & Idade <= 10 ~ "<= 10",
                      Idade > 10 & Idade <= 20 ~ "10-20",
                      Idade > 20 ~ "> 20")) %>%
  group_by(Ano, Tipo, region, ff_area, idade) %>% 
  summarise(
    Temp = mean(Temp),
    n_obs = n(),
    .groups = "drop"
  )

all_lulc_temp_stages <- all_lulc_temp %>%
  mutate(Tipo = case_when(Tipo == "PF" ~ "PF",
                          Tipo == "SF" & idade == "<= 10" ~ "SF_early",
                          Tipo == "SF" & idade == "10-20" ~ "SF_intermediate",
                          Tipo == "SF" & idade == "> 20" ~ "SF_advanced",
                          TRUE ~ NA_character_ )) %>% 

  filter(!is.na(Tipo))  


all_lulc_temp_final <- all_lulc_temp_stages %>% 
  mutate(type = case_when(Tipo == "PF" ~ 1,
                          Tipo == "SF_early" ~ 2, 
                          Tipo == "SF_intermediate" ~ 3,
                          Tipo == "SF_advanced" ~ 4))

max_min_temp = all_lulc_temp_final %>%
  group_by(Tipo, region, ff_area, idade) %>%
  summarise(
    max_temp = max(Temp),
    median_temp = median(Temp),
    min_temp = min(Temp)
  )

write_xlsx(max_min_temp, "data_group/plan_max_and_min_group_temp.xlsx")

write_xlsx(all_lulc_temp_final, "data_group/plan_data_group_temp.xlsx")

rm(all_lulc_50_final)

gc()
```

### Tests for Normality, Homoscedasticity, and Independence

```{r}

all_pf_sf_temp = read_xlsx("data_group/plan_data_group_temp.xlsx")

shapiro_test_results_temp <- all_pf_sf_temp %>%
  group_by(Tipo,region, ff_area) %>%
  summarise(
    p_value = shapiro.test(Temp)$p.value
  ) %>%
  ungroup()

print(shapiro_test_results_temp)

write_xlsx(shapiro_test_results_temp, "data_group/shapiro_test_temp_results.xlsx")


all_pf_sf_temp$Tipo <- as.factor(all_pf_sf_temp$Tipo)

results_list_temp <- list()

for (region in unique(all_pf_sf_temp$region)) {
  for (ff_area in unique(all_pf_sf_temp$ff_area)) {

    data_subset <- all_pf_sf_temp %>%
      filter(region == !!region, ff_area == !!ff_area)

    if (nrow(data_subset) > 1 && length(unique(data_subset$Tipo)) > 1) {
      try({
        levene_test_temp <- leveneTest(Temp ~ Tipo, data = data_subset)
        p_value <- levene_test_temp$`Pr(>F)`[1]  
      }, silent = TRUE)
    } else {
      p_value <- NA  
    }

    if (!is.null(p_value)) {
      results_list_temp[[length(results_list_temp) + 1]] <- data.frame(region = region, 
                                                             ff_area = ff_area, 
                                                             p_value = p_value)
    }
  }
}

levene_test_results_temp <- do.call(rbind, results_list_temp)

print(levene_test_results_temp)

write_xlsx(levene_test_results_temp, "data_group/levene_test_temp_results.xlsx")


kendall_results_list_temp <- list()

for (region in unique(all_pf_sf_temp$region)) {
  for (ff_area in unique(all_pf_sf_temp$ff_area)) {

    data_subset <- all_pf_sf_temp %>%
      filter(region == !!region, ff_area == !!ff_area)

    if (nrow(data_subset) > 1 && length(unique(data_subset$Tipo)) > 1) {
      try({
        cor_test <- cor.test(as.numeric(data_subset$Temp), as.numeric(data_subset$Tipo), method = "kendall")
        p_value <- cor_test$p.value  
        cor_value <- cor_test$estimate  
      }, silent = TRUE)
    } else {
      p_value <- NA  
      cor_value <- NA
    }

    if (!is.null(p_value)) {
      kendall_results_list_temp[[length(kendall_results_list_temp) + 1]] <- data.frame(
        region = region, 
        ff_area = ff_area, 
        cor_value = cor_value,
        p_value = p_value
      )
    }
  }
}

kendall_test_results_temp <- do.call(rbind, kendall_results_list_temp)

write_xlsx(kendall_test_results_temp, "data_group/kendall_test_temp_results.xlsx")

gc()

```

### t-test between coverage types – PF and SF

```{r}

area_ranges <- c("1-10", "10-20", "20-30", "30-40", "40-50", 
                 "50-60", "60-70", "70-80", "80-90", "90-100")

regions <- c("CBA", "EBA", "SBA")

significance_results_t_temp <- data.frame(Intervalo = area_ranges)  

for (region in regions) {
  
  p_values_pf_sf_early <- c()
  p_values_pf_sf_intermediate <- c()
  p_values_pf_sf_advanced <- c()

  for (range in area_ranges) {

    pv_data <- all_pf_sf_temp %>% filter(region == !!region, ff_area == range, Tipo == "PF")
    sv_early_data <- all_pf_sf_temp %>% filter(region == !!region, ff_area == range, Tipo == "SF_early")
    sv_intermediate_data <- all_pf_sf_temp %>% filter(region == !!region, ff_area == range, Tipo == "SF_intermediate")
    sv_advanced_data <- all_pf_sf_temp %>% filter(region == !!region, ff_area == range, Tipo == "SF_advanced")

    if (nrow(pv_data) > 1 && nrow(sv_early_data) > 1) {
      p_values_pf_sf_early <- c(p_values_pf_sf_early, t.test(pv_data$Temp, sv_early_data$Temp, var.equal = TRUE)$p.value)
    } else {
      p_values_pf_sf_early <- c(p_values_pf_sf_early, NA)
    }
    
    if (nrow(pv_data) > 1 && nrow(sv_intermediate_data) > 1) {
      p_values_pf_sf_intermediate <- c(p_values_pf_sf_intermediate, t.test(pv_data$Temp, sv_intermediate_data$Temp, var.equal = TRUE)$p.value)
    } else {
      p_values_pf_sf_intermediate <- c(p_values_pf_sf_intermediate, NA)
    }
    
    if (nrow(pv_data) > 1 && nrow(sv_advanced_data) > 1) {
      p_values_pf_sf_advanced <- c(p_values_pf_sf_advanced, t.test(pv_data$Temp, sv_advanced_data$Temp, var.equal = TRUE)$p.value)
    } else {
      p_values_pf_sf_advanced <- c(p_values_pf_sf_advanced, NA)
    }
  }

  significance_results_t_temp[[paste0(region, "_PF_vs_SF_early")]] <- p_values_pf_sf_early
  significance_results_t_temp[[paste0(region, "_PF_vs_SF_intermediate")]] <- p_values_pf_sf_intermediate
  significance_results_t_temp[[paste0(region, "_PF_vs_SF_advanced")]] <- p_values_pf_sf_advanced
}

write_xlsx(significance_results_t_temp, "data_group/significance_results_temp_testt.xlsx")

gc()

```

### Wilcoxon test between coverage types – PF and SF

```{r}

all_pf_sf_temp = read_xlsx("data_group/plan_data_group_temp.xlsx")

area_ranges <- c("1-10", "10-20", "20-30", "30-40", "40-50", 
                 "50-60", "60-70", "70-80", "80-90", "90-100")

regions <- c("CBA", "EBA", "SBA")

significance_results_w_temp <- data.frame(Intervalo = area_ranges)  

for (region in regions) {
  
  p_values_pf_sf_early <- c()
  p_values_pf_sf_intermediate <- c()
  p_values_pf_sf_advanced <- c()

  for (range in area_ranges) {

    pv_data <- all_pf_sf_temp %>% filter(region == !!region, ff_area == range, Tipo == "PF")
    sv_early_data <- all_pf_sf_temp %>% filter(region == !!region, ff_area == range, Tipo == "SF_early")
    sv_intermediate_data <- all_pf_sf_temp %>% filter(region == !!region, ff_area == range, Tipo == "SF_intermediate")
    sv_advanced_data <- all_pf_sf_temp %>% filter(region == !!region, ff_area == range, Tipo == "SF_advanced")

    if (nrow(pv_data) > 1 && nrow(sv_early_data) > 1) {
      p_values_pf_sf_early <- c(p_values_pf_sf_early, wilcox.test(pv_data$Temp, sv_early_data$Temp, paired = FALSE)$p.value)
    } else {
      p_values_pf_sf_early <- c(p_values_pf_sf_early, NA)
    }
    
    if (nrow(pv_data) > 1 && nrow(sv_intermediate_data) > 1) {
      p_values_pf_sf_intermediate <- c(p_values_pf_sf_intermediate, wilcox.test(pv_data$Temp, sv_intermediate_data$Temp, paired = FALSE)$p.value)
    } else {
      p_values_pf_sf_intermediate <- c(p_values_pf_sf_intermediate, NA)
    }
    
    if (nrow(pv_data) > 1 && nrow(sv_advanced_data) > 1) {
      p_values_pf_sf_advanced <- c(p_values_pf_sf_advanced, wilcox.test(pv_data$Temp, sv_advanced_data$Temp, paired = FALSE)$p.value)
    } else {
      p_values_pf_sf_advanced <- c(p_values_pf_sf_advanced, NA)
    }
  }

  significance_results_w_temp[[paste0(region, "_PF_vs_SF_early")]] <- p_values_pf_sf_early
  significance_results_w_temp[[paste0(region, "_PF_vs_SF_intermediate")]] <- p_values_pf_sf_intermediate
  significance_results_w_temp[[paste0(region, "_PF_vs_SF_advanced")]] <- p_values_pf_sf_advanced
}

write_xlsx(significance_results_w_temp, "data_group/significance_results_temp_Wilcox_test.xlsx")

gc()
```

#### Plot with significance

```{r}

long_significance_results_w_temp <- significance_results_w_temp %>%
  gather(key = "Comparison", value = "P_value", -Intervalo) %>%
  separate(Comparison, into = c("region", "Comparison_Type"), sep = "_PF_vs_")

all_pf_sf_temp <- all_pf_sf_temp %>%
  left_join(long_significance_results_w_temp, 
            by = c("ff_area" = "Intervalo", "region" = "region", "Tipo" = "Comparison_Type"))

all_pf_sf_temp$Significant <- ifelse(all_pf_sf_temp$P_value < 0.05, "*", "")

all_pf_sf_temp <- all_pf_sf_temp %>%
  mutate(Tipo = factor(Tipo, levels = c("PF", "SF_early", "SF_intermediate", "SF_advanced")))

get_whisker_max <- function(x) {
  stats <- boxplot.stats(x)$stats
  return(stats[5])
}


significance_positions <- all_pf_sf_temp %>%
  group_by(ff_area, region, Tipo) %>%
  summarise(max_whisker = get_whisker_max(Temp), .groups = "drop") %>%
  left_join(long_significance_results_w_temp, 
            by = c("ff_area" = "Intervalo", "region" = "region", "Tipo" = "Comparison_Type")) %>%
  filter(P_value < 0.05) %>%
  mutate(
    asterisk_y = max_whisker + 0.1, 
    x_adjust = case_when(
      Tipo == "SF_early" ~ -0.1,
      Tipo == "SF_intermediate" ~ 0.1,
      Tipo == "SF_advanced" ~ 0.3
    ),
    x_pos = as.numeric(factor(ff_area, levels = area_ranges)) + x_adjust
  )

p_ast_temp_w <-all_pf_sf_temp %>% 
  ggplot(aes(x=ff_area, y=Temp, fill=Tipo)) +
  geom_boxplot(outlier.alpha = 0.4) +
  facet_grid(~region) + 
  ggthemes::theme_few() +
  theme(
    legend.position = c(0.91, 0.84),
    legend.background = element_rect(fill="transparent"),
    text = element_text(size = 12, face = "bold",family = "serif"), 
    #text = element_text(size = 7, face = "bold",family = "serif"), 
    axis.text = element_text(size = 10, face = "plain", family = "serif", color = "black"),
    #axis.text.y = element_text(size = 12, face = "plain", family = "serif", color = "black"),
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
    legend.text = element_text(size = 9),
    #legend.text = element_text(size = 7),
    panel.spacing.x = unit(1, "lines"),  
    panel.spacing.y = unit(1, "lines"),
    axis.title = element_text(size = 14, margin = margin(r = 20)),
    strip.text = element_blank()) +
  scale_fill_manual(values = c("#7F3B08", "#FEE0B6", "#FDB863", "#E08214"), name = NULL, labels = c("PF", "SF early", "SF intermediate", "SF advanced")) +
  scale_y_continuous(breaks = 23:27, limits = c(NA, 27), expand = expansion(mult = c(0.01, 0.05)), 
  labels = label_number(width = 6, zero.print = TRUE)) + 
   labs(x = "Fractional cover (%)",
        y = "Average annual mean \n temperature (°C)",
        #y = expression(atop("Average annual mean", paste("temperature (", degree, "C·yr"^-1, ")")))
        ) +
  geom_text(data = significance_positions, 
            aes(x = x_pos, y = asterisk_y, label = "*"),
            size = 3, color = "red")

p_ast_temp_w

ggsave("fig/area_x_temp_pf_sf_significant_asterisco_wilcoxon.png", dpi = 600, width = 9, height = 5.5)

gc()

```

## Panel all variables

```{r}

p_ret_prec_w <- p_ast_prec_w + 
  scale_x_discrete(expand = c(0.05, 0.05)) + 
  theme(plot.margin = margin(t = 10, r = 10, b = 5, l = 20))  

p_ret_evapo_w <- p_ast_evapo_w + 
  scale_x_discrete(expand = c(0.05, 0.05)) + 
  theme(plot.margin = margin(t = 5, r = 10, b = 5, l = 20))   

p_ret_temp_w <- p_ast_temp_w + 
  scale_x_discrete(expand = c(0.05, 0.05)) + 
  theme(plot.margin = margin(t = 5, r = 10, b = 10, l = 20))  


painel_obs <- ggarrange(
  p_ret_prec_w, p_ret_evapo_w, p_ret_temp_w,
  ncol = 1, nrow = 3,
  labels = c("A", "B", "C"),  
  font.label = list(color = "black", face = "bold", size = 14, family = "serif"),
  align = "v",                
  heights = c(1, 1, 1)       
)

ggsave("fig/painels/painel_fraction_forest_x_prec.png", 
       plot = painel_obs, dpi = 600, width = 10, height = 12, bg = "white")

ggsave("fig/painels/painel_fraction_forest_x_prec.pdf", 
       plot = painel_obs, width = 10, height = 12, bg = "white")

```

# Modeling

## Functions

```{r}

safe_dir_create <- function(path) {
  dir.create(path, showWarnings = FALSE, recursive = TRUE)
}

safe_quantile <- function(x, p) {
  if (all(is.na(x))) return(NA_real_)
  as.numeric(quantile(x, p, na.rm = TRUE))
}

calc_metrics <- function(observed, fitted) {
  
  ok <- is.finite(observed) & is.finite(fitted)
  observed <- observed[ok]
  fitted <- fitted[ok]
  
  if (length(observed) < 5) {
    return(data.table(
      CCC = NA_real_,
      RMSE = NA_real_,
      MAE = NA_real_,
      Bias = NA_real_,
      R2 = NA_real_,
      Residual_SD = NA_real_,
      n_eval = length(observed)
    ))
  }
  
  resid <- observed - fitted
  
  ccc_val <- tryCatch(
    DescTools::CCC(observed, fitted)$rho.c[, "est"],
    error = function(e) NA_real_
  )
  
  data.table(
    CCC = as.numeric(ccc_val),
    RMSE = sqrt(mean(resid^2, na.rm = TRUE)),
    MAE = mean(abs(resid), na.rm = TRUE),
    Bias = mean(fitted - observed, na.rm = TRUE),
    R2 = cor(observed, fitted, use = "complete.obs")^2,
    Residual_SD = sd(resid, na.rm = TRUE),
    n_eval = length(observed)
  )
}

sample_cells_region <- function(cell_table,
                                region_name,
                                n_cells_per_stratum = 250,
                                seed = NULL) {
  
  if (!is.null(seed)) set.seed(seed)
  
  ct <- as.data.table(cell_table)[region == region_name]
  
  selected <- ct[
    ,
    .(cell_id = sample(cell_id, size = min(.N, n_cells_per_stratum))),
    by = .(Tipo_cell, frac_class)
  ]
  
  unique(selected$cell_id)
}

extract_year_effects <- function(model, region_name, run_id, response_var) {
  
  ran <- ranef(model)
  year_name <- grep("Ano", names(ran), value = TRUE)[1]
  
  if (is.na(year_name)) return(NULL)
  
  year_eff <- as.data.frame(ran[[year_name]])
  year_eff$Ano <- rownames(year_eff)
  names(year_eff)[1] <- "year_effect"
  
  year_eff %>%
    mutate(
      region = region_name,
      response = response_var,
      run = run_id,
      Ano = as.numeric(as.character(Ano)),
      year_effect = as.numeric(year_effect)
    )
}

make_spatial_component_fast <- function(model,
                                        dados_sub,
                                        fitted_full,
                                        fixed_part,
                                        region_name,
                                        run_id,
                                        response_var,
                                        max_points = 6000) {
  
  ran <- ranef(model)
  year_name <- grep("Ano", names(ran), value = TRUE)[1]
  
  if (is.na(year_name)) {
    dados_sub$year_effect <- 0
  } else {
    year_eff_df <- as.data.frame(ran[[year_name]])
    year_eff_df$Ano <- rownames(year_eff_df)
    names(year_eff_df)[1] <- "year_effect"
    
    year_eff_df <- year_eff_df %>%
      mutate(
        Ano = as.character(Ano),
        year_effect = as.numeric(year_effect)
      )
    
    dados_sub <- dados_sub %>%
      mutate(Ano_chr = as.character(Ano)) %>%
      left_join(year_eff_df, by = c("Ano_chr" = "Ano")) %>%
      mutate(year_effect = ifelse(is.na(year_effect), 0, year_effect))
  }
  
  tmp <- dados_sub %>%
    mutate(
      fitted_full = fitted_full,
      fixed_part = fixed_part,
      spatial_component = fitted_full - fixed_part - year_effect
    )
  
  spatial_cell <- tmp %>%
    group_by(cell_id, Lon, Lat) %>%
    summarise(
      spatial_component = mean(spatial_component, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    mutate(
      region = region_name,
      response = response_var,
      run = run_id
    )
  
  if (nrow(spatial_cell) > max_points) {
    spatial_cell <- spatial_cell %>%
      slice_sample(n = max_points)
  }
  
  spatial_cell
}

run_one_spamm_loaded <- function(region_name,
                                 run_id,
                                 data_region,
                                 cell_table,
                                 out_dir,
                                 response_var,
                                 n_cells_per_stratum = 250,
                                 df_frac = 3,
                                 seed = NULL,
                                 save_model = TRUE,
                                 save_run_data = TRUE,
                                 diagnostic_sample_max = 30000) {
  
  spaMM.options(nb_cores = 1)
  
  run_chr <- sprintf("%03d", run_id)
  
  file_done <- file.path(out_dir, paste0("DONE_run_", run_chr, "_", region_name, ".txt"))
  file_model <- file.path(out_dir, "models", paste0("model_run_", run_chr, "_", region_name, ".rds"))
  file_data <- file.path(out_dir, "run_data", paste0("data_run_", run_chr, "_", region_name, ".rds"))
  file_contrasts <- file.path(out_dir, "contrasts", paste0("contrasts_run_", run_chr, "_", region_name, ".csv"))
  file_year <- file.path(out_dir, "year_effects", paste0("year_effects_run_", run_chr, "_", region_name, ".csv"))
  file_spatial <- file.path(out_dir, "spatial_component", paste0("spatial_component_run_", run_chr, "_", region_name, ".csv"))
  file_diag <- file.path(out_dir, "diagnostics", paste0("diagnostics_run_", run_chr, "_", region_name, ".csv"))
  file_metrics <- file.path(out_dir, "metrics", paste0("metrics_run_", run_chr, "_", region_name, ".csv"))
  file_log <- file.path(out_dir, "logs", paste0("log_run_", run_chr, "_", region_name, ".csv"))
  
  if (
    file.exists(file_done) &&
    file.exists(file_contrasts) &&
    file.exists(file_year) &&
    file.exists(file_spatial) &&
    file.exists(file_diag) &&
    file.exists(file_metrics)
  ) {
    message("Skipping completed run ", run_id, " | ", region_name)
    return(NULL)
  }
  
  message("\n===================================")
  message("Running run ", run_id, " | region ", region_name)
  message("===================================")
  
  t0 <- Sys.time()
  
  log_status <- "success"
  log_error <- NA_character_
  
  result <- tryCatch({
    
    selected_cells <- sample_cells_region(
      cell_table = cell_table,
      region_name = region_name,
      n_cells_per_stratum = n_cells_per_stratum,
      seed = seed
    )
    
    dados_sub <- as.data.table(data_region)[
      cell_id %in% selected_cells,
      .(
        response,
        cell_id,
        Ano,
        Tipo,
        frac_area,
        X_km_s,
        Y_km_s,
        Lon,
        Lat
      )
    ]
    
    dados_sub <- droplevels(as.data.frame(dados_sub))
    
    form <- as.formula(
      paste0(
        "response ~ Tipo * ns(frac_area, df = ", df_frac, ") + ",
        "(1 | Ano) + ",
        "Matern(1 | X_km_s + Y_km_s)"
      )
    )
    
    model <- fitme(
      form,
      data = dados_sub,
      family = gaussian(),
      method = "REML"
    )
    
    if (save_model) {
      saveRDS(model, file_model)
    }
    
    if (save_run_data) {
      saveRDS(dados_sub, file_data)
    }
    
    fixed_form <- as.formula(
      paste0("~ Tipo * ns(frac_area, df = ", df_frac, ")")
    )
    
    beta <- fixef(model)
    
    
    # Spatially adjusted contrasts
    frac_seq <- seq(0.01, 1.00, length.out = 50)
    
    newdata_adj <- expand.grid(
      frac_area = frac_seq,
      Tipo = levels(dados_sub$Tipo),
      KEEP.OUT.ATTRS = FALSE,
      stringsAsFactors = FALSE
    )
    
    newdata_adj$Tipo <- factor(
      newdata_adj$Tipo,
      levels = levels(dados_sub$Tipo)
    )
    
    X_adj <- model.matrix(fixed_form, data = newdata_adj)
    common_adj <- intersect(colnames(X_adj), names(beta))
    
    newdata_adj$pred <- as.numeric(
      X_adj[, common_adj, drop = FALSE] %*% beta[common_adj]
    )
    
    contrasts_adj <- newdata_adj %>%
      select(frac_area, Tipo, pred) %>%
      pivot_wider(names_from = Tipo, values_from = pred) %>%
      mutate(
        region = region_name,
        response = response_var,
        run = run_id,
        contrast_type = "Spatially adjusted",
        PF_minus_SF_early = PF - SF_early,
        PF_minus_SF_intermediate = PF - SF_intermediate,
        PF_minus_SF_advanced = PF - SF_advanced
      ) %>%
      select(
        region, response, run, contrast_type, frac_area,
        PF_minus_SF_early,
        PF_minus_SF_intermediate,
        PF_minus_SF_advanced
      )
    

    # Full prediction, residuals and diagnostics
    fitted_full <- as.numeric(
      predict(model, newdata = dados_sub, type = "response")
    )
    
    residual_model <- dados_sub$response - fitted_full
    
    diagnostics <- data.table(
      region = region_name,
      response = response_var,
      run = run_id,
      observed = dados_sub$response,
      fitted = fitted_full,
      residual = residual_model,
      Tipo = as.character(dados_sub$Tipo),
      frac_area = dados_sub$frac_area,
      Ano = as.character(dados_sub$Ano),
      cell_id = dados_sub$cell_id,
      Lon = dados_sub$Lon,
      Lat = dados_sub$Lat
    )
    
    if (nrow(diagnostics) > diagnostic_sample_max) {
      set.seed(seed + 999)
      diagnostics_save <- diagnostics[
        sample(.N, diagnostic_sample_max)
      ]
    } else {
      diagnostics_save <- diagnostics
    }
    
    fwrite(diagnostics_save, file_diag)
    
    metrics <- calc_metrics(
      observed = diagnostics$observed,
      fitted = diagnostics$fitted
    )
    
    metrics[, `:=`(
      region = region_name,
      response = response_var,
      run = run_id,
      n_cells_per_stratum = n_cells_per_stratum,
      selected_cells = length(selected_cells),
      rows_model = nrow(dados_sub),
      n_unique_cells = dplyr::n_distinct(dados_sub$cell_id)
    )]
    
    setcolorder(
      metrics,
      c(
        "region", "response", "run",
        "CCC", "RMSE", "MAE", "Bias", "R2", "Residual_SD", "n_eval",
        "n_cells_per_stratum", "selected_cells", "rows_model", "n_unique_cells"
      )
    )
    
    fwrite(metrics, file_metrics)
    

    # Current landscape contrasts
    dados_pred <- dados_sub %>%
      mutate(
        pred_full = fitted_full,
        frac_area_bin = cut(
          frac_area,
          breaks = seq(0, 1, by = 0.05),
          include.lowest = TRUE
        )
      )
    
    landscape_mean <- dados_pred %>%
      group_by(frac_area_bin, Tipo) %>%
      summarise(
        frac_area = median(frac_area, na.rm = TRUE),
        pred = mean(pred_full, na.rm = TRUE),
        .groups = "drop"
      )
    
    pf_landscape <- landscape_mean %>%
      filter(Tipo == "PF") %>%
      dplyr::select(frac_area_bin, frac_area, PF = pred)
    
    sf_landscape <- landscape_mean %>%
      filter(Tipo != "PF") %>%
      dplyr::select(frac_area_bin, Tipo, pred) %>%
      pivot_wider(names_from = Tipo, values_from = pred)
    
    for (nm in c("SF_early", "SF_intermediate", "SF_advanced")) {
      if (!nm %in% names(sf_landscape)) sf_landscape[[nm]] <- NA_real_
    }
    
    contrasts_landscape <- pf_landscape %>%
      left_join(sf_landscape, by = "frac_area_bin") %>%
      mutate(
        region = region_name,
        response = response_var,
        run = run_id,
        contrast_type = "Current landscape",
        PF_minus_SF_early = PF - SF_early,
        PF_minus_SF_intermediate = PF - SF_intermediate,
        PF_minus_SF_advanced = PF - SF_advanced
      ) %>%
      dplyr::select(
        region, response, run, contrast_type, frac_area,
        PF_minus_SF_early,
        PF_minus_SF_intermediate,
        PF_minus_SF_advanced
      )
    
    contrasts <- bind_rows(
      contrasts_adj,
      contrasts_landscape
    )
    
    fwrite(contrasts, file_contrasts)
    

    # Year and spatial components
    year_effects <- extract_year_effects(
      model = model,
      region_name = region_name,
      run_id = run_id,
      response_var = response_var
    )
    
    fwrite(year_effects, file_year)
    
    X_sub <- model.matrix(fixed_form, data = dados_sub)
    common_sub <- intersect(colnames(X_sub), names(beta))
    
    fixed_part <- as.numeric(
      X_sub[, common_sub, drop = FALSE] %*% beta[common_sub]
    )
    
    spatial_component <- make_spatial_component_fast(
      model = model,
      dados_sub = dados_sub,
      fitted_full = fitted_full,
      fixed_part = fixed_part,
      region_name = region_name,
      run_id = run_id,
      response_var = response_var,
      max_points = 6000
    )
    
    fwrite(spatial_component, file_spatial)
    
    writeLines("done", file_done)
    
    invisible(TRUE)
    
  }, error = function(e) {
    
    log_status <<- "error"
    log_error <<- conditionMessage(e)
    message("ERROR in run ", run_id, " | ", region_name, ": ", log_error)
    invisible(FALSE)
  })
  
  elapsed_min <- round(
    as.numeric(difftime(Sys.time(), t0, units = "mins")),
    2
  )
  
  fwrite(
    data.table(
      region = region_name,
      response = response_var,
      run = run_id,
      status = log_status,
      error_message = log_error,
      n_cells_per_stratum = n_cells_per_stratum,
      df_frac = df_frac,
      elapsed_min = elapsed_min
    ),
    file_log
  )
  
  message("Finished run ", run_id, " | status: ", log_status, " | ", elapsed_min, " min")
  
  invisible(gc())
  NULL
}

```

## PRECIPITATION

### Parameters

```{r,eval = FALSE, echo = TRUE}

spaMM.options(save.HLfit = FALSE)
spaMM.options(nb_cores = 1)

set.seed(123)

regions <- c("CBA") #("CBA", "EBA", "SBA")
response_var <- "Prec"

n_runs <- 30 
n_cells_per_stratum <- 200
df_frac <- 3
n_workers <- 4

data_dir <- paste0("data_by_region_", tolower(response_var))
out_dir <- paste0("spamm_ensemble_", tolower(response_var), "_diagnostics_CBA")

safe_dir_create(data_dir)
safe_dir_create(out_dir)

for (subdir in c(
  "models", "run_data", "contrasts", "year_effects",
  "spatial_component", "diagnostics", "metrics", "logs", "figures", "tables"
)) {
  safe_dir_create(file.path(out_dir, subdir))
}

```

### Data

```{r,eval = FALSE, echo = TRUE}

file_in <- "/media/amazonia/hd2/lais/GitHub/paper_climate_forest_regeneration/data_group/plan_all_lulc_final.csv"

dt <- fread(file_in)

dt <- dt[Tipo %in% c("PF", "SF")]

dt[, cell_id := .GRP, by = .(Lon, Lat)]

dt <- dt %>%
  transmute(
    cell_id = as.integer(cell_id),
    response = .data[[response_var]],
    frac_area = frac_area,
    Tipo_raw = Tipo,
    Ano = factor(Ano),
    region = Region,
    Idade = Idade_SF,
    Lon = Lon,
    Lat = Lat
  ) %>%
  mutate(
    idade = case_when(
      Idade >= 0  & Idade <= 10 ~ "<= 10",
      Idade > 10 & Idade <= 20 ~ "10-20",
      Idade > 20 ~ "> 20",
      TRUE ~ NA_character_
    ),
    Tipo = case_when(
      Tipo_raw == "PF" ~ "PF",
      Tipo_raw == "SF" & idade == "<= 10" ~ "SF_early",
      Tipo_raw == "SF" & idade == "10-20" ~ "SF_intermediate",
      Tipo_raw == "SF" & idade == "> 20" ~ "SF_advanced",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(
    !is.na(response),
    !is.na(frac_area),
    !is.na(Tipo),
    !is.na(Ano),
    !is.na(Lon),
    !is.na(Lat),
    frac_area >= 0.01,
    frac_area <= 1.00
  ) %>%
  mutate(
    Tipo = factor(
      Tipo,
      levels = c("PF", "SF_early", "SF_intermediate", "SF_advanced")
    ),
    Tipo_cell = as.character(Tipo),
    frac_class = cut(
      frac_area,
      breaks = seq(0, 1, by = 0.20),
      include.lowest = TRUE,
      right = TRUE
    )
  )

setDT(dt)

stratum_year_summary <- dt[
  ,
  .(n_years = uniqueN(Ano)),
  by = .(region, Tipo_cell, frac_class, cell_id)
][
  ,
  .(
    n_cells_total = .N,
    n_cells_21_years = sum(n_years == 21),
    prop_cells_21_years = mean(n_years == 21),
    max_years = max(n_years)
  ),
  by = .(region, Tipo_cell, frac_class)
][order(region, Tipo_cell, frac_class)]

fwrite(
  stratum_year_summary,
  file.path(out_dir, "tables", "num_pixels_estratos.csv")
)

for (r in regions) {
  
  dt_r <- copy(dt[region == r])
  
  mean_lat <- mean(dt_r$Lat, na.rm = TRUE)
  
  dt_r[, X_km := Lon * 111 * cos(mean_lat * pi / 180)]
  dt_r[, Y_km := Lat * 111]
  dt_r[, X_km_s := X_km - mean(X_km, na.rm = TRUE)]
  dt_r[, Y_km_s := Y_km - mean(Y_km, na.rm = TRUE)]
  
  dt_r <- dt_r[
    ,
    .(
      response,
      cell_id,
      Ano,
      Tipo,
      frac_area,
      X_km_s,
      Y_km_s,
      Lon,
      Lat,
      region,
      Tipo_cell,
      frac_class
    )
  ]
  
  saveRDS(
    dt_r,
    file.path(data_dir, paste0("data_", r, ".rds"))
  )
}

cell_table_all <- unique(
  dt[
    ,
    .(region, Tipo_cell, frac_class, cell_id)
  ]
)

saveRDS(
  cell_table_all,
  file.path(data_dir, "cell_table_all.rds")
)

rm(dt, stratum_year_summary, cell_table_all)
invisible(gc())

```

### Model

```{r,eval = FALSE, echo = TRUE}

run_ids <- seq_len(n_runs)

if (n_workers == 1) {
  
  region_name <- regions[1]
  
  data_region_local <- readRDS(
    file.path(data_dir, paste0("data_", region_name, ".rds"))
  )
  
  cell_table_local <- readRDS(
    file.path(data_dir, "cell_table_all.rds")
  )
  
  for (i in run_ids) {
    run_one_spamm_loaded(
      region_name = region_name,
      run_id = i,
      data_region = data_region_local,
      cell_table = cell_table_local,
      out_dir = out_dir,
      response_var = response_var,
      n_cells_per_stratum = n_cells_per_stratum,
      df_frac = df_frac,
      seed = 1000 + i,
      save_model = TRUE,
      save_run_data = TRUE
    )
    invisible(gc())
  }
  
} else {
  
  plan(multisession, workers = n_workers)
  
  run_chunks <- split(
    run_ids,
    rep(seq_len(n_workers), length.out = length(run_ids))
  )
  
  future_lapply(
    run_chunks,
    function(run_vec) {
      
      library(data.table)
      library(dplyr)
      library(tidyr)
      library(spaMM)
      library(splines)
      library(DescTools)
      
      spaMM.options(save.HLfit = FALSE)
      spaMM.options(nb_cores = 1)
      
      region_name <- regions[1]
      
      data_region_local <- readRDS(
        file.path(data_dir, paste0("data_", region_name, ".rds"))
      )
      
      cell_table_local <- readRDS(
        file.path(data_dir, "cell_table_all.rds")
      )
      
      for (i in run_vec) {
        run_one_spamm_loaded(
          region_name = region_name,
          run_id = i,
          data_region = data_region_local,
          cell_table = cell_table_local,
          out_dir = out_dir,
          response_var = response_var,
          n_cells_per_stratum = n_cells_per_stratum,
          df_frac = df_frac,
          seed = 1000 + i,
          save_model = TRUE,
          save_run_data = TRUE
        )
        invisible(gc())
      }
      
      NULL
    },
    future.seed = TRUE
  )
  
  plan(sequential)
}

```

## EVAPOTRANSPIRATION

### Parameters

```{r,eval = FALSE, echo = TRUE}

spaMM.options(save.HLfit = FALSE)
spaMM.options(nb_cores = 1)

set.seed(123)

regions <- c("SBA") #("CBA", "EBA", "SBA")
response_var <- "Evapo"

n_runs <- 30 
n_cells_per_stratum <- 200
df_frac <- 3
n_workers <- 4

data_dir <- paste0("data_by_region_", tolower(response_var))
out_dir <- paste0("spamm_ensemble_", tolower(response_var), "_diagnostics_SBA")

safe_dir_create(data_dir)
safe_dir_create(out_dir)

for (subdir in c(
  "models", "run_data", "contrasts", "year_effects",
  "spatial_component", "diagnostics", "metrics", "logs", "figures", "tables"
)) {
  safe_dir_create(file.path(out_dir, subdir))
}

```

### Data

```{r,eval = FALSE, echo = TRUE}

file_in <- "/media/amazonia/hd2/lais/GitHub/paper_climate_forest_regeneration/data_group/plan_all_lulc_final.csv"

dt <- fread(file_in)

dt <- dt[Tipo %in% c("PF", "SF")]

dt[, cell_id := .GRP, by = .(Lon, Lat)]

dt <- dt %>%
  filter(!Evapo >= 2000) %>% 
  transmute(
    cell_id = as.integer(cell_id),
    response = .data[[response_var]],
    frac_area = frac_area,
    Tipo_raw = Tipo,
    Ano = factor(Ano),
    region = Region,
    Idade = Idade_SF,
    Lon = Lon,
    Lat = Lat
  ) %>%
  mutate(
    idade = case_when(
      Idade >= 0  & Idade <= 10 ~ "<= 10",
      Idade > 10 & Idade <= 20 ~ "10-20",
      Idade > 20 ~ "> 20",
      TRUE ~ NA_character_
    ),
    Tipo = case_when(
      Tipo_raw == "PF" ~ "PF",
      Tipo_raw == "SF" & idade == "<= 10" ~ "SF_early",
      Tipo_raw == "SF" & idade == "10-20" ~ "SF_intermediate",
      Tipo_raw == "SF" & idade == "> 20" ~ "SF_advanced",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(
    !is.na(response),
    !is.na(frac_area),
    !is.na(Tipo),
    !is.na(Ano),
    !is.na(Lon),
    !is.na(Lat),
    frac_area >= 0.01,
    frac_area <= 1.00
  ) %>%
  mutate(
    Tipo = factor(
      Tipo,
      levels = c("PF", "SF_early", "SF_intermediate", "SF_advanced")
    ),
    Tipo_cell = as.character(Tipo),
    frac_class = cut(
      frac_area,
      breaks = seq(0, 1, by = 0.20),
      include.lowest = TRUE,
      right = TRUE
    )
  )

setDT(dt)

stratum_year_summary <- dt[
  ,
  .(n_years = uniqueN(Ano)),
  by = .(region, Tipo_cell, frac_class, cell_id)
][
  ,
  .(
    n_cells_total = .N,
    n_cells_21_years = sum(n_years == 21),
    prop_cells_21_years = mean(n_years == 21),
    max_years = max(n_years)
  ),
  by = .(region, Tipo_cell, frac_class)
][order(region, Tipo_cell, frac_class)]

fwrite(
  stratum_year_summary,
  file.path(out_dir, "tables", "num_pixels_estratos.csv")
)

for (r in regions) {
  
  dt_r <- copy(dt[region == r])
  
  mean_lat <- mean(dt_r$Lat, na.rm = TRUE)
  
  dt_r[, X_km := Lon * 111 * cos(mean_lat * pi / 180)]
  dt_r[, Y_km := Lat * 111]
  dt_r[, X_km_s := X_km - mean(X_km, na.rm = TRUE)]
  dt_r[, Y_km_s := Y_km - mean(Y_km, na.rm = TRUE)]
  
  dt_r <- dt_r[
    ,
    .(
      response,
      cell_id,
      Ano,
      Tipo,
      frac_area,
      X_km_s,
      Y_km_s,
      Lon,
      Lat,
      region,
      Tipo_cell,
      frac_class
    )
  ]
  
  saveRDS(
    dt_r,
    file.path(data_dir, paste0("data_", r, ".rds"))
  )
}

cell_table_all <- unique(
  dt[
    ,
    .(region, Tipo_cell, frac_class, cell_id)
  ]
)

saveRDS(
  cell_table_all,
  file.path(data_dir, "cell_table_all.rds")
)

rm(dt, stratum_year_summary, cell_table_all)
invisible(gc())

```

### Model

```{r,eval = FALSE, echo = TRUE}

run_ids <- seq_len(n_runs)

if (n_workers == 1) {
  
  region_name <- regions[1]
  
  data_region_local <- readRDS(
    file.path(data_dir, paste0("data_", region_name, ".rds"))
  )
  
  cell_table_local <- readRDS(
    file.path(data_dir, "cell_table_all.rds")
  )
  
  for (i in run_ids) {
    run_one_spamm_loaded(
      region_name = region_name,
      run_id = i,
      data_region = data_region_local,
      cell_table = cell_table_local,
      out_dir = out_dir,
      response_var = response_var,
      n_cells_per_stratum = n_cells_per_stratum,
      df_frac = df_frac,
      seed = 1000 + i,
      save_model = TRUE,
      save_run_data = TRUE
    )
    invisible(gc())
  }
  
} else {
  
  plan(multisession, workers = n_workers)
  
  run_chunks <- split(
    run_ids,
    rep(seq_len(n_workers), length.out = length(run_ids))
  )
  
  future_lapply(
    run_chunks,
    function(run_vec) {
      
      library(data.table)
      library(dplyr)
      library(tidyr)
      library(spaMM)
      library(splines)
      library(DescTools)
      
      spaMM.options(save.HLfit = FALSE)
      spaMM.options(nb_cores = 1)
      
      region_name <- regions[1]
      
      data_region_local <- readRDS(
        file.path(data_dir, paste0("data_", region_name, ".rds"))
      )
      
      cell_table_local <- readRDS(
        file.path(data_dir, "cell_table_all.rds")
      )
      
      for (i in run_vec) {
        run_one_spamm_loaded(
          region_name = region_name,
          run_id = i,
          data_region = data_region_local,
          cell_table = cell_table_local,
          out_dir = out_dir,
          response_var = response_var,
          n_cells_per_stratum = n_cells_per_stratum,
          df_frac = df_frac,
          seed = 1000 + i,
          save_model = TRUE,
          save_run_data = TRUE
        )
        invisible(gc())
      }
      
      NULL
    },
    future.seed = TRUE
  )
  
  plan(sequential)
}

```

##TEMPERATURE

### Parameters

```{r,eval = FALSE, echo = TRUE}

spaMM.options(save.HLfit = FALSE)
spaMM.options(nb_cores = 1)

set.seed(123)

regions <-  c("CBA") #("CBA", "EBA", "SBA")
response_var <- "Temp"

n_runs <- 30 
n_cells_per_stratum <- 200
df_frac <- 3
n_workers <- 4

data_dir <- paste0("data_by_region_", tolower(response_var))
out_dir <- paste0("spamm_ensemble_", tolower(response_var), "_diagnostics_CBA")

safe_dir_create(data_dir)
safe_dir_create(out_dir)

for (subdir in c(
  "models", "run_data", "contrasts", "year_effects",
  "spatial_component", "diagnostics", "metrics", "logs", "figures", "tables"
)) {
  safe_dir_create(file.path(out_dir, subdir))
}

```

### Data

```{r,eval = FALSE, echo = TRUE}

file_in <- "/media/amazonia/hd2/lais/GitHub/paper_climate_forest_regeneration/data_group/plan_all_lulc_final.csv"

dt <- fread(file_in)

dt <- dt[Tipo %in% c("PF", "SF")]

dt[, cell_id := .GRP, by = .(Lon, Lat)]

dt <- dt %>%
  transmute(
    cell_id = as.integer(cell_id),
    response = .data[[response_var]],
    frac_area = frac_area,
    Tipo_raw = Tipo,
    Ano = factor(Ano),
    region = Region,
    Idade = Idade_SF,
    Lon = Lon,
    Lat = Lat
  ) %>%
  mutate(
    idade = case_when(
      Idade >= 0  & Idade <= 10 ~ "<= 10",
      Idade > 10 & Idade <= 20 ~ "10-20",
      Idade > 20 ~ "> 20",
      TRUE ~ NA_character_
    ),
    Tipo = case_when(
      Tipo_raw == "PF" ~ "PF",
      Tipo_raw == "SF" & idade == "<= 10" ~ "SF_early",
      Tipo_raw == "SF" & idade == "10-20" ~ "SF_intermediate",
      Tipo_raw == "SF" & idade == "> 20" ~ "SF_advanced",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(
    !is.na(response),
    !is.na(frac_area),
    !is.na(Tipo),
    !is.na(Ano),
    !is.na(Lon),
    !is.na(Lat),
    frac_area >= 0.01,
    frac_area <= 1.00
  ) %>%
  mutate(
    Tipo = factor(
      Tipo,
      levels = c("PF", "SF_early", "SF_intermediate", "SF_advanced")
    ),
    Tipo_cell = as.character(Tipo),
    frac_class = cut(
      frac_area,
      breaks = seq(0, 1, by = 0.20),
      include.lowest = TRUE,
      right = TRUE
    )
  )

setDT(dt)

stratum_year_summary <- dt[
  ,
  .(n_years = uniqueN(Ano)),
  by = .(region, Tipo_cell, frac_class, cell_id)
][
  ,
  .(
    n_cells_total = .N,
    n_cells_21_years = sum(n_years == 21),
    prop_cells_21_years = mean(n_years == 21),
    max_years = max(n_years)
  ),
  by = .(region, Tipo_cell, frac_class)
][order(region, Tipo_cell, frac_class)]

fwrite(
  stratum_year_summary,
  file.path(out_dir, "tables", "num_pixels_estratos.csv")
)

for (r in regions) {
  
  dt_r <- copy(dt[region == r])
  
  mean_lat <- mean(dt_r$Lat, na.rm = TRUE)
  
  dt_r[, X_km := Lon * 111 * cos(mean_lat * pi / 180)]
  dt_r[, Y_km := Lat * 111]
  dt_r[, X_km_s := X_km - mean(X_km, na.rm = TRUE)]
  dt_r[, Y_km_s := Y_km - mean(Y_km, na.rm = TRUE)]
  
  dt_r <- dt_r[
    ,
    .(
      response,
      cell_id,
      Ano,
      Tipo,
      frac_area,
      X_km_s,
      Y_km_s,
      Lon,
      Lat,
      region,
      Tipo_cell,
      frac_class
    )
  ]
  
  saveRDS(
    dt_r,
    file.path(data_dir, paste0("data_", r, ".rds"))
  )
}

cell_table_all <- unique(
  dt[
    ,
    .(region, Tipo_cell, frac_class, cell_id)
  ]
)

saveRDS(
  cell_table_all,
  file.path(data_dir, "cell_table_all.rds")
)

rm(dt, stratum_year_summary, cell_table_all)
invisible(gc())

```

### Model

```{r,eval = FALSE, echo = TRUE}

run_ids <- seq_len(n_runs)

if (n_workers == 1) {
  
  region_name <- regions[1]
  
  data_region_local <- readRDS(
    file.path(data_dir, paste0("data_", region_name, ".rds"))
  )
  
  cell_table_local <- readRDS(
    file.path(data_dir, "cell_table_all.rds")
  )
  
  for (i in run_ids) {
    run_one_spamm_loaded(
      region_name = region_name,
      run_id = i,
      data_region = data_region_local,
      cell_table = cell_table_local,
      out_dir = out_dir,
      response_var = response_var,
      n_cells_per_stratum = n_cells_per_stratum,
      df_frac = df_frac,
      seed = 1000 + i,
      save_model = TRUE,
      save_run_data = TRUE
    )
    invisible(gc())
  }
  
} else {
  
  plan(multisession, workers = n_workers)
  
  run_chunks <- split(
    run_ids,
    rep(seq_len(n_workers), length.out = length(run_ids))
  )
  
  future_lapply(
    run_chunks,
    function(run_vec) {
      
      library(data.table)
      library(dplyr)
      library(tidyr)
      library(spaMM)
      library(splines)
      library(DescTools)
      
      spaMM.options(save.HLfit = FALSE)
      spaMM.options(nb_cores = 1)
      
      region_name <- regions[1]
      
      data_region_local <- readRDS(
        file.path(data_dir, paste0("data_", region_name, ".rds"))
      )
      
      cell_table_local <- readRDS(
        file.path(data_dir, "cell_table_all.rds")
      )
      
      for (i in run_vec) {
        run_one_spamm_loaded(
          region_name = region_name,
          run_id = i,
          data_region = data_region_local,
          cell_table = cell_table_local,
          out_dir = out_dir,
          response_var = response_var,
          n_cells_per_stratum = n_cells_per_stratum,
          df_frac = df_frac,
          seed = 1000 + i,
          save_model = TRUE,
          save_run_data = TRUE
        )
        invisible(gc())
      }
      
      NULL
    },
    future.seed = TRUE
  )
  
  plan(sequential)
}

```

# Figures

## CBA

### Main figure 

#### Base functions

```{r}

safe_quantile <- function(x, p) {
  if (all(is.na(x))) return(NA_real_)
  as.numeric(quantile(x, p, na.rm = TRUE))
}

read_all_files <- function(path, pattern) {
  files <- list.files(path, pattern = pattern, full.names = TRUE)
  if (length(files) == 0) return(data.table())
  rbindlist(lapply(files, fread), fill = TRUE)
}

get_out_dir <- function(response_var, region_name) {
  paste0("spamm_ensemble_", tolower(response_var), "_diagnostics_", region_name)
}


region_name <- "CBA"
response_vars <- c("Prec", "Evapo", "Temp")

response_labels <- c(
  Prec = "Precipitation",
  Evapo   = "Evapotranspiration",
  Temp  = "Land surface temperature")

response_units <- c(
  Prec = "mm",
  Evapo   = "mm",
  Temp  = "°C")

cols_by_response <- list(
  Prec = c(
    "SF[e]" = "#87CEFA",
    "SF[i]" = "#1E90FF",
    "SF[a]" = "#0000FF"),
  Evapo = c(
    "SF[e]" = "lightgreen",
    "SF[i]" = "green1",
    "SF[a]" = "green4"),
  Temp = c(
    "SF[e]" = "#FEE0B6",
    "SF[i]" = "#FDB863",
    "SF[a]" = "#E08214"))

color_year <- c(
  Prec  = "midnightblue",
  Evapo = "#02401B",
  Temp  = "#7F3B08")

x_forest_scale <- scale_x_continuous(
  limits = c(0, 1),
  breaks = seq(0, 1, by = 0.25),
  labels = scales::percent_format(accuracy = 1),
  expand = expansion(mult = c(0.01, 0.01))
)

rdBu_10 <- c(
  "#B2182B", "#D6604D", "#F4A582", "#FDDBC7", "#F7F7F7",
  "#D1E5F0", "#92C5DE", "#4393C3", "#2166AC", "#053061"
)

rdBu_10_temp <- c("#053061", "#2166AC","#4393C3", "#92C5DE", "#D1E5F0",
                  "#F7F7F7", "#FDDBC7", "#F4A582", "#D6604D", "#B2182B")

bin_levels <- c(
      "0–5", "5–10", "10–15", "15–20", "20–25",
      "25–30", "30–35", "35–40", "40–45", "45–50",
      "50–55", "55–60", "60–65", "65–70", "70–75",
      "75–80", "80–85", "85–90", "90–95", "95–100")
 
load_response_results <- function(response_var, region_name) {
  
  out_dir <- get_out_dir(response_var, region_name)
  
  contrasts_ensemble <- read_all_files(
  file.path(out_dir, "contrasts"),
  "^contrasts_run_.*\\.csv$"
  )
  
  year_effects_ensemble <- read_all_files(
    file.path(out_dir, "year_effects"),
    "^year_effects_run_.*\\.csv$"
  )
  
  spatial_component_ensemble <- read_all_files(
    file.path(out_dir, "spatial_component"),
    "^spatial_component_run_.*\\.csv$"
  )
  
  diagnostics_ensemble <- read_all_files(
    file.path(out_dir, "diagnostics"),
    "^diagnostics_run_.*\\.csv$"
  )
  
  metrics_ensemble <- read_all_files(
    file.path(out_dir, "metrics"),
    "^metrics_run_.*\\.csv$"
  )
  
  logs_ensemble <- read_all_files(
    file.path(out_dir, "logs"),
    "^log_run_.*\\.csv$"
  )
  
  fwrite(contrasts_ensemble, file.path(out_dir, "tables", "contrasts_ensemble_all_runs.csv"),  sep = ";",  dec = ",")
  fwrite(year_effects_ensemble, file.path(out_dir, "tables", "year_effects_ensemble_all_runs.csv"),  sep = ";",  dec = ",")
  fwrite(spatial_component_ensemble, file.path(out_dir, "tables", "spatial_component_ensemble_all_runs.csv"),  sep = ";",  dec = ",")
  fwrite(diagnostics_ensemble, file.path(out_dir, "tables", "diagnostics_ensemble_sampled_all_runs.csv"),  sep = ";",  dec = ",")
  fwrite(metrics_ensemble, file.path(out_dir, "tables", "metrics_ensemble_all_runs.csv"),  sep = ";",  dec = ",")
  fwrite(logs_ensemble, file.path(out_dir, "tables", "logs_ensemble_all_runs.csv"),  sep = ";",  dec = ",")
  
  contrasts_long <- contrasts_ensemble %>%
  pivot_longer(
    cols = starts_with("PF_minus_SF"),
    names_to = "stage",
    values_to = "difference"
  ) %>%
  mutate(
    stage = recode(
      stage,
      "PF_minus_SF_early" = "SF[e]",
      "PF_minus_SF_intermediate" = "SF[i]",
      "PF_minus_SF_advanced" = "SF[a]"
    ),
    stage = factor(stage, levels = c("SF[e]", "SF[i]", "SF[a]")),
    contrast_type = factor(
      contrast_type,
      levels = c("Spatially adjusted", "Current landscape")
    ),
    frac_bin = case_when(
      contrast_type == "Current landscape" ~ as.character(cut(
        frac_area,
        breaks = seq(0, 1, by = 0.05),
        labels = bin_levels,
        include.lowest = TRUE,
        right = TRUE
      )),
      TRUE ~ NA_character_
    ),
    frac_bin = factor(frac_bin, levels = bin_levels),
    frac_plot = case_when(
      contrast_type == "Spatially adjusted" ~ round(frac_area, 6),
      contrast_type == "Current landscape" ~ as.numeric(frac_bin)
    ),
    response = response_var
  )
  
  contrasts_summary <- contrasts_long %>%
    filter(!is.na(frac_plot), !is.na(difference)) %>%
    group_by(region, response, contrast_type, frac_plot, frac_bin, stage) %>%
    summarise(
      median = median(difference, na.rm = TRUE),
      lower = safe_quantile(difference, 0.025),
      upper = safe_quantile(difference, 0.975),
      n_runs = n_distinct(run),
      .groups = "drop"
    )
  
  fwrite(
  contrasts_summary,
  file.path(out_dir, "tables", "contrasts_summary_empirical_CI.csv"),  sep = ";",  dec = ",")
  
  year_summary <- year_effects_ensemble %>%
    mutate(response = response_var) %>%
    group_by(region, response, Ano) %>%
    summarise(
      median = median(year_effect, na.rm = TRUE),
      lower = safe_quantile(year_effect, 0.025),
      upper = safe_quantile(year_effect, 0.975),
      n_runs = sum(!is.na(year_effect)),
      .groups = "drop"
    )
  
  fwrite(
  year_summary,
  file.path(out_dir, "tables", "year_effects_summary_empirical_CI.csv"),  sep = ";",  dec = ",")
  
  spatial_summary <- spatial_component_ensemble %>%
    mutate(response = response_var) %>%
    group_by(region, response, Lon, Lat) %>%
    summarise(
      spatial_median = median(spatial_component, na.rm = TRUE),
      spatial_mean = mean(spatial_component, na.rm = TRUE),
      spatial_sd = sd(spatial_component, na.rm = TRUE),
      n_runs = n(),
      .groups = "drop"
    )
  
  fwrite(
  spatial_summary,
  file.path(out_dir, "tables", "spatial_component_summary.csv"),  sep = ";",  dec = ",")
  
  metrics_summary <- metrics_ensemble %>%
    group_by(region, response) %>%
    summarise(
      n_successful_runs = n(),
      
      CCC_median = median(CCC, na.rm = TRUE),
      CCC_lower = safe_quantile(CCC, 0.025),
      CCC_upper = safe_quantile(CCC, 0.975),
      
      RMSE_median = median(RMSE, na.rm = TRUE),
      RMSE_lower = safe_quantile(RMSE, 0.025),
      RMSE_upper = safe_quantile(RMSE, 0.975),
      
      MAE_median = median(MAE, na.rm = TRUE),
      MAE_lower = safe_quantile(MAE, 0.025),
      MAE_upper = safe_quantile(MAE, 0.975),
      
      Bias_median = median(Bias, na.rm = TRUE),
      Bias_lower = safe_quantile(Bias, 0.025),
      Bias_upper = safe_quantile(Bias, 0.975),
      
      R2_median = median(R2, na.rm = TRUE),
      R2_lower = safe_quantile(R2, 0.025),
      R2_upper = safe_quantile(R2, 0.975),
      
      .groups = "drop"
    )

  fwrite(
    metrics_summary,
    file.path(out_dir, "tables", "metrics_summary_empirical_CI.csv"),  sep = ";",  dec = ",")

  
  list(
    contrasts_summary = contrasts_summary,
    year_summary = year_summary,
    spatial_summary = spatial_summary
  )
}

```

#### Plots functions

```{r}

spatial_breaks <- list(
  Prec  = seq(-750, 750, by = 250),
  Evapo = seq(-150, 150, by = 50),
  Temp  = seq(-1.5, 1.5, by = 0.5)
)

spatial_labels <- list(
  Prec  = c("≤ -750", "-500", "-250", "0", "250", "500", "≥ 750"),
  Evapo = c("≤ -150", "-100", "-50", "0", "50", "100", "≥ 150"),
  Temp  = c("≤ -1.5", "-1", "-0.5", "0", "0.5", "1", "≥ 1.5")
)

spatial_cols <- list(
  Prec  = rdBu_10,
  Evapo = rdBu_10,
  Temp  = rdBu_10_temp
)

make_p_adj <- function(dat, response_var, show_legend = FALSE, first_col = FALSE) {
  
  unit <- response_units[[response_var]]
  label <- response_labels[[response_var]]
  
  ggplot(
    dat %>% filter(contrast_type == "Spatially adjusted"),
    aes(x = frac_plot, y = median, color = stage, fill = stage)
  ) +
    geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.35) +
    geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.18, color = NA) +
    geom_line(linewidth = 0.8) +
    x_forest_scale +
    scale_color_manual(
      values = cols_by_response[[response_var]],
      labels = parse(text = c("bold(SF[e])", "bold(SF[i])", "bold(SF[a])"))) +
    scale_fill_manual(
      values = cols_by_response[[response_var]],
      labels = parse(text = c("bold(SF[e])", "bold(SF[i])", "bold(SF[a])")),
      guide = "none") +
    labs(
      title = label,
      x = NULL,
      y = ifelse(
        first_col,
        paste0("Spatially adjusted\nPF − SF(", unit, ")"),
        paste0("\nPF − SF(", unit, ")")),
      color = NULL,
      fill = NULL
    ) +
    theme_bw(base_size = 10) +
    theme(
      text = element_text(face = "bold", family = "serif"),
      plot.title = element_text(hjust = 0.5, size = 12),
      axis.title = element_text(size = 10),
      axis.text = element_text(size = 9, color = "black"),
      legend.text = element_text(size = 9, face = "bold"),
      legend.key.width = unit(0.45, "cm"),
      legend.key.height = unit(0.25, "cm"),
      legend.margin = margin(t = -3, r = 0, b = -6, l = 0),
      legend.box.margin = margin(t = -6, r = 0, b = -8, l = 0),
      plot.margin = margin(t = 2, r = 6, b = 10, l = 4),
      legend.position = ifelse(show_legend, "top", "none"),
      legend.direction = "horizontal",
      panel.grid.minor = element_blank()
    )
}

make_p_land <- function(dat, response_var, show_legend = FALSE, first_col = FALSE) {
  
  unit <- response_units[[response_var]]
  
  ggplot(
    dat %>% filter(contrast_type == "Current landscape"),
    aes(x = frac_bin, y = median, color = stage, group = stage)
  ) +
    geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.35) +
    geom_line(linewidth = 0.8, linetype = "22") +
    scale_x_discrete(
      drop = FALSE,
      breaks = bin_levels[c(FALSE, TRUE)]) +
    scale_color_manual(
      values = cols_by_response[[response_var]],
      labels = parse(text = c("bold(SF[e])", "bold(SF[i])", "bold(SF[a])"))
    ) +
    labs(
      x = "Forest cover (%)",
      y = ifelse(
        first_col,
        paste0("Current landscape\nPF − SF(", unit, ")"),
        paste0("\nPF − SF(", unit, ")")
      ),
      color = NULL
    ) +
    theme_bw(base_size = 10) +
    theme(
      text = element_text(face = "bold", family = "serif"),
      axis.title = element_text(size = 10),
      axis.text.y = element_text(size = 9, color = "black"),
      axis.text.x = element_text(size = 9, color = "black", angle = 45, hjust = 1),
      legend.text = element_text(size = 9, face = "bold"),
      legend.key.width = unit(0.45, "cm"),
      legend.key.height = unit(0.25, "cm"),
      legend.margin = margin(t = -3, r = 0, b = -6, l = 0),
      legend.box.margin = margin(t = -6, r = 0, b = -8, l = 0),
      plot.margin = margin(t = 2, r = 6, b = 10, l = 4),
      legend.position = ifelse(show_legend, "top", "none"),
      legend.direction = "horizontal",
      panel.grid.minor = element_blank()
    )
}

make_p_year <- function(dat, response_var, first_col = FALSE) {
  
  unit <- response_units[[response_var]]
  
  ggplot(dat, aes(x = Ano, y = median)) +
    geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.35) +
    geom_ribbon(aes(ymin = lower, ymax = upper), fill = color_year[[response_var]], alpha = 0.18) +
    geom_line(linewidth = 0.85, color = color_year[[response_var]]) +
    geom_point(size = 1.5, color = color_year[[response_var]]) +
    scale_x_continuous(
      breaks = seq(2001, 2021, by = 5),
      limits = c(2001, 2021)
    ) +
    labs(
      x = NULL,
      y = ifelse(
        first_col,
        paste0("Year effect\n(", unit, ")"),
        paste0("\n(", unit, ")"))) +
    theme_bw(base_size = 10) +
    theme(
      text = element_text(face = "bold", family = "serif"),
      axis.title = element_text(size = 10),
      axis.text = element_text(size = 9, color = "black"),
      plot.margin = margin(t = 2, r = 4, b = 10, l = 4),
      panel.grid.minor = element_blank()
    )
}

make_p_spatial <- function(dat, response_var, first_col = FALSE) {
  
  unit <- response_units[[response_var]]
  
  ggplot(dat, aes(x = Lon, y = Lat, z = spatial_median)) +
    stat_summary_2d(fun = median, bins = 120) +
    coord_quickmap(expand = TRUE) +
    scale_fill_stepsn(
      colours = spatial_cols[[response_var]],
      breaks = spatial_breaks[[response_var]],
      labels = spatial_labels[[response_var]],
      limits = range(spatial_breaks[[response_var]]),
      oob = scales::squish,
      name = paste0("Spatial deviation (", unit, ")"),
      guide = guide_coloursteps(
        title.position = "top",
        title.hjust = 0.5,
        label.position = "bottom",
        barwidth = unit(5.0, "cm"),
        barheight = unit(0.32, "cm"))) +
    labs(x = "Longitude", 
         y = ifelse(first_col, "Latitude", "")) +
    theme_bw(base_size = 10) +
    theme(
      text = element_text(face = "bold", family = "serif"),
      axis.title = element_text(size = 10),
      axis.text = element_text(size = 9, color = "black"),
      legend.position = "top",
      legend.direction = "horizontal",
      legend.title = element_text(size = 9, face = "bold"),
      legend.text = element_text(size = 8.5),
      legend.box.margin = margin(t = -4, r = 0, b = -2, l = 0),
      legend.margin = margin(t = -4, r = 0, b = -4, l = 0),
      plot.margin = margin(t = 2, r = 2, b = 0, l = 2),
      panel.grid.minor = element_blank())
  
}

```

#### Results

```{r}

all_results <- lapply(response_vars, load_response_results, region_name = region_name)
names(all_results) <- response_vars


p_adj_list <- Map(
  function(v, fc)
    make_p_adj(
      all_results[[v]]$contrasts_summary,
      response_var = v,
      show_legend = TRUE,
      first_col = fc
    ),
  response_vars,
  c(TRUE, FALSE, FALSE)
)

p_land_list <- Map(
  function(v, fc)
    make_p_land(
      all_results[[v]]$contrasts_summary,
      response_var = v,
      show_legend = TRUE,
      first_col = fc
    ),
  response_vars,
  c(TRUE, FALSE, FALSE)
)

p_year_list <- Map(
  function(v, fc)
    make_p_year(
      all_results[[v]]$year_summary,
      response_var = v,
      first_col = fc
    ),
  response_vars,
  c(TRUE, FALSE, FALSE)
)

p_spatial_list <- Map(
  function(v, fc)
    make_p_spatial(
      all_results[[v]]$spatial_summary,
      response_var = v,
      first_col = fc
    ),
  response_vars,
  c(TRUE, FALSE, FALSE)
)

out_fig_dir <- file.path("fig")

p_adj_list
ggsave(
  file.path(out_fig_dir, paste0("figure_spat_adj_", region_name, ".png")),
  p_adj_list, width = 9, height = 5.5, dpi = 600, bg = "white")

p_land_list
ggsave(
  file.path(out_fig_dir, paste0("figure_curr_land_", region_name, ".png")),
  p_land_list, width = 9, height = 5.5, dpi = 600, bg = "white")

p_year_list
ggsave(
  file.path(out_fig_dir, paste0("figure_year_effect_", region_name, ".png")),
  p_year_list, width = 9, height = 5.5, dpi = 600, bg = "white")

p_spatial_list
ggsave(
  file.path(out_fig_dir, paste0("figure_spat_effect_", region_name, ".png")),
  p_spatial_list, width = 9, height = 5.5, dpi = 600, bg = "white")


```

#### Panel

```{r}

tag_plot <- function(p, tag) {
  p + labs(tag = tag) +
    theme(
      plot.tag = element_text(face = "bold", size = 13, family = "serif"),
      plot.tag.position = c(0, 1)
    )
}

p_adj_list <- Map(tag_plot, p_adj_list, c("A", "E", "I"))
p_land_list <- Map(tag_plot, p_land_list, c("B", "F", "J"))
p_year_list <- Map(tag_plot, p_year_list, c("C", "G", "K"))
p_spatial_list <- Map(tag_plot, p_spatial_list, c("D", "H", "L"))

fig_region <- 
  wrap_plots(p_adj_list, nrow = 1) /
  wrap_plots(p_land_list, nrow = 1) /
  wrap_plots(p_year_list, nrow = 1) /
  wrap_plots(p_spatial_list, nrow = 1,  widths = c(1, 1, 1)) +
  plot_layout(heights = c(1, 1, 0.9, 1.20)) +
  plot_annotation(
    title = region_name,
    theme = theme(
      plot.title = element_text(
        size = 16,
        face = "bold",
        family = "serif",
        hjust = 0.5
      )
    )
  ) &
  theme(
    text = element_text(face = "bold", family = "serif"),
    plot.tag = element_text(face = "bold", size = 13)
  )

fig_region

out_fig_dir <- file.path("fig")
dir.create(out_fig_dir, showWarnings = FALSE, recursive = TRUE)

ggsave(
  file.path(out_fig_dir, paste0("painels/figure_ensemble_", region_name, "_all_variables.png")),
  fig_region, width = 10, height = 12, dpi = 600, bg = "white"
)

ggsave(file.path(out_fig_dir, paste0("painels/figure_ensemble_", region_name, "_all_variables.pdf")), 
       plot = fig_region, width = 10, height = 12, bg = "white")

```

### Diagnostics 

#### Functions

```{r}

safe_quantile <- function(x, p) {
  if (all(is.na(x))) return(NA_real_)
  as.numeric(quantile(x, p, na.rm = TRUE))
}

read_all_files <- function(path, pattern) {
  files <- list.files(path, pattern = pattern, full.names = TRUE)
  if (length(files) == 0) return(data.table())
  rbindlist(lapply(files, fread), fill = TRUE)
}

get_out_dir <- function(response_var, region_name) {
  paste0("spamm_ensemble_", tolower(response_var), "_diagnostics_", region_name)
}


region_name <- "CBA"
response_vars <- c("Prec", "Evapo", "Temp")

response_labels <- c(
  Prec  = "Precipitation",
  Evapo = "Evapotranspiration",
  Temp  = "Land surface temperature"
)

response_units <- c(
  Prec  = "mm",
  Evapo = "mm",
  Temp  = "°C"
)

color_year <- c(
  Prec  = "midnightblue",
  Evapo = "#02401B",
  Temp  = "#7F3B08"
)


load_diag_results <- function(response_var, region_name) {
  
  out_dir <- get_out_dir(response_var, region_name)
  
  diagnostics_ensemble <- read_all_files(
    file.path(out_dir, "diagnostics"),
    "^diagnostics_run_.*\\.csv$"
  )
  
  metrics_ensemble <- read_all_files(
    file.path(out_dir, "metrics"),
    "^metrics_run_.*\\.csv$"
  )
  
  if (nrow(diagnostics_ensemble) == 0) {
    stop("No diagnostics files found for ", response_var, " in ", out_dir)
  }
  
  if (nrow(metrics_ensemble) == 0) {
    stop("No metrics files found for ", response_var, " in ", out_dir)
  }
  
  diagnostics_ensemble <- diagnostics_ensemble %>%
    mutate(
      response = response_var,
      variable_label = response_labels[[response_var]]
    )
  
  metrics_ensemble <- metrics_ensemble %>%
    mutate(
      response = response_var,
      variable_label = response_labels[[response_var]]
    )
  
  list(
    diagnostics = diagnostics_ensemble,
    metrics = metrics_ensemble
  )
}


theme_diag <- function() {
  theme_bw(base_size = 11) +
    theme(
      text = element_text(face = "bold", family = "serif"),
      plot.title = element_text(hjust = 0.5, size = 12),
      axis.title = element_text(size = 11),
      axis.text = element_text(size = 10, color = "black"),
      strip.text = element_blank(),
      strip.background = element_blank(),
      panel.grid.minor = element_blank(),
      aspect.ratio = 1,
      plot.margin = margin(t = 3, r = 5, b = 4, l = 5)
    )
}



sample_diag <- function(dat, n_max = 30000) {
  set.seed(123)
  as.data.table(dat)[
    ,
    .SD[sample(.N, min(.N, n_max))],
    by = response
  ]
}

```

#### Plots

```{r}

make_p_resid <- function(dat, response_var, first_col = FALSE) {
  
  unit <- response_units[[response_var]]
  label <- response_labels[[response_var]]
  col <- color_year[[response_var]]
  
  ggplot(dat, aes(x = fitted, y = residual)) +
    geom_point(alpha = 0.14, size = 0.22, color = col) +
    geom_smooth(method = "loess", se = FALSE, color = "black", linewidth = 0.55) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "grey35", linewidth = 0.35) +
    labs(
      title = label,
      x = paste0("Fitted (", unit, ")"),
      y = ifelse(first_col, paste0("Residuals\n(", unit, ")"), paste0("\n(", unit, ")"))
    ) +
    theme_diag()
}

make_p_qq <- function(dat, response_var, first_col = FALSE) {
  
  unit <- response_units[[response_var]]
  col <- color_year[[response_var]]
  
  ggplot(dat, aes(sample = residual)) +
    stat_qq(size = 0.25, alpha = 0.25, color = col) +
    stat_qq_line(linewidth = 0.55, color = "black") +
    labs(
      x = "Theoretical quantiles",
      y = ifelse( first_col,  paste0("Sample quantiles\n(", unit, ")"),  paste0("\n(", unit, ")"))) +
    theme_diag()
}

make_p_obs_pred <- function(dat, metrics, response_var, first_col = FALSE) {
  
  unit <- response_units[[response_var]]
  col <- color_year[[response_var]]
  
  ccc_med <- median(metrics$CCC, na.rm = TRUE)
  r2_med <- median(metrics$R2, na.rm = TRUE)
  
  label_txt <- paste0(
    "CCC = ", round(ccc_med, 2),
    "\nR² = ", round(r2_med, 2)
  )
  
  ggplot(dat, aes(x = observed, y = fitted)) +
    geom_point(alpha = 0.18, size = 0.25, color = col) +
    geom_smooth(method = "lm", se = FALSE, color = "black", linewidth = 0.55) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey35", linewidth = 0.35) +
    annotate(
      "text",
      x = -Inf, y = Inf,
      label = label_txt,
      hjust = -0.1, vjust = 1.5,
      size = 3.5,
      family = "serif",
      fontface = "bold"
    ) +
    labs(
      x = paste0("Observed (", unit, ")"),
      y = ifelse(first_col, paste0("Predicted\n(", unit, ")"), paste0("\n(", unit, ")"))
    ) +
    theme_diag()
}

make_p_accuracy <- function(metrics, response_var, first_col = FALSE) {
  
  col <- color_year[[response_var]]
  
  dat <- metrics %>%
    select(region, run, CCC, R2) %>%
    pivot_longer(
      cols = c(CCC, R2),
      names_to = "metric",
      values_to = "value"
    ) %>%
    mutate(
      metric = factor(metric, levels = c("CCC", "R2"), labels = c("CCC", "R²"))
    )
  
  ggplot(dat, aes(x = metric, y = value)) +
    geom_boxplot(
      width = 0.45,
      outlier.shape = NA,
      linewidth = 0.45,
      fill = "white",
      color = col
    ) +
    geom_jitter(
      width = 0.08,
      alpha = 0.65,
      size = 1.2,
      color = col
    ) +
    stat_summary(
      fun = median,
      geom = "point",
      shape = 23,
      size = 2.4,
      fill = "white",
      color = "black"
    ) +
    coord_cartesian(ylim = c(0, 1)) +
    scale_y_continuous(breaks = seq(0, 1, by = 0.25)) +
    labs(
      x = NULL,
      y = ifelse(first_col, "Agreement /\nexplained variance", "")
    ) +
    theme_diag()
}

make_p_error <- function(metrics, response_var, first_col = FALSE) {
  
  unit <- response_units[[response_var]]
  col <- color_year[[response_var]]
  
  dat <- metrics %>%
    select(region, run, RMSE, MAE) %>%
    pivot_longer(
      cols = c(RMSE, MAE),
      names_to = "metric",
      values_to = "value"
    ) %>%
    mutate(
      metric = factor(metric, levels = c("RMSE", "MAE"))
    )
  
  ggplot(dat, aes(x = metric, y = value)) +
    geom_boxplot(
      width = 0.45,
      outlier.shape = NA,
      linewidth = 0.45,
      fill = "white",
      color = col
    ) +
    geom_jitter(
      width = 0.08,
      alpha = 0.65,
      size = 1.2,
      color = col
    ) +
    stat_summary(
      fun = median,
      geom = "point",
      shape = 23,
      size = 2.4,
      fill = "white",
      color = "black"
    ) +
    labs(
      x = NULL,
      y = ifelse(first_col, paste0("Error metrics\n(", unit, ")"), paste0("\n(", unit, ")"))
    ) +
    theme_diag()
}

```

#### Panel

```{r}

diag_results <- lapply(response_vars, load_diag_results, region_name = region_name)
names(diag_results) <- response_vars

diag_sampled <- lapply(diag_results, function(x) {
  sample_diag(x$diagnostics, n_max = 30000)
})

metrics_list <- lapply(diag_results, function(x) {
  x$metrics
})


first_col_flags <- c(TRUE, FALSE, FALSE)

p_resid_list <- Map(
  function(v, fc) {
    make_p_resid(
      diag_sampled[[v]],
      response_var = v,
      first_col = fc
    )
  },
  response_vars,
  first_col_flags
)

p_qq_list <- Map(
  function(v, fc) {
    make_p_qq(
      diag_sampled[[v]],
      response_var = v,
      first_col = fc
    )
  },
  response_vars,
  first_col_flags
)

p_obs_list <- Map(
  function(v, fc) {
    make_p_obs_pred(
      diag_sampled[[v]],
      metrics = metrics_list[[v]],
      response_var = v,
      first_col = fc
    )
  },
  response_vars,
  first_col_flags
)

p_acc_list <- Map(
  function(v, fc) {
    make_p_accuracy(
      metrics_list[[v]],
      response_var = v,
      first_col = fc
    )
  },
  response_vars,
  first_col_flags
)

p_err_list <- Map(
  function(v, fc) {
    make_p_error(
      metrics_list[[v]],
      response_var = v,
      first_col = fc
    )
  },
  response_vars,
  first_col_flags
)


tag_plot <- function(p, tag) {
  p + labs(tag = tag) +
    theme(
      plot.tag = element_text(face = "bold", size = 13, family = "serif"),
      plot.tag.position = c(0, 1)
    )
}

p_resid_list <- Map(tag_plot, p_resid_list, c("A", "E", "I"))
p_qq_list    <- Map(tag_plot, p_qq_list,    c("B", "F", "J"))
p_obs_list   <- Map(tag_plot, p_obs_list,   c("C", "G", "K"))
p_err_list   <- Map(tag_plot, p_err_list,   c("D", "H", "L"))


fig_diag_region <-
  wrap_plots(
    c(
      p_resid_list,
      p_qq_list,
      p_obs_list,
      p_err_list
    ),
    ncol = 3,
    byrow = TRUE
  ) +
  plot_layout(
    widths = c(1, 1, 1),
    heights = c(1, 1, 1, 1)
  ) +
  plot_annotation(
    title = paste0(region_name),
    theme = theme(
      plot.title = element_text(
        size = 16,
        face = "bold",
        family = "serif",
        hjust = 0.5
      )
    )
  ) &
  theme(
    text = element_text(face = "bold", family = "serif"),
    plot.tag = element_text(face = "bold", size = 13)
  )

fig_diag_region

ggsave(
  file.path(out_fig_dir, paste0("painels/figure_diagnostics_", region_name, "_all_variables.png")),
  fig_diag_region, width = 10, height = 11.5, dpi = 600, bg = "white")

ggsave(file.path(out_fig_dir, paste0("painels/figure_diagnostics_", region_name, "_all_variables.pdf")), 
       plot = fig_diag_region, width = 10, height = 11.5, bg = "white")

```

## EBA

### Main figure 

#### Base functions

```{r}

safe_quantile <- function(x, p) {
  if (all(is.na(x))) return(NA_real_)
  as.numeric(quantile(x, p, na.rm = TRUE))
}

read_all_files <- function(path, pattern) {
  files <- list.files(path, pattern = pattern, full.names = TRUE)
  if (length(files) == 0) return(data.table())
  rbindlist(lapply(files, fread), fill = TRUE)
}

get_out_dir <- function(response_var, region_name) {
  paste0("spamm_ensemble_", tolower(response_var), "_diagnostics_", region_name)
}


region_name <- "EBA"
response_vars <- c("Prec", "Evapo", "Temp")

response_labels <- c(
  Prec = "Precipitation",
  Evapo   = "Evapotranspiration",
  Temp  = "Land surface temperature")

response_units <- c(
  Prec = "mm",
  Evapo   = "mm",
  Temp  = "°C")

cols_by_response <- list(
  Prec = c(
    "SF[e]" = "#87CEFA",
    "SF[i]" = "#1E90FF",
    "SF[a]" = "#0000FF"),
  Evapo = c(
    "SF[e]" = "lightgreen",
    "SF[i]" = "green1",
    "SF[a]" = "green4"),
  Temp = c(
    "SF[e]" = "#FEE0B6",
    "SF[i]" = "#FDB863",
    "SF[a]" = "#E08214"))

color_year <- c(
  Prec  = "midnightblue",
  Evapo = "#02401B",
  Temp  = "#7F3B08")

x_forest_scale <- scale_x_continuous(
  limits = c(0, 1),
  breaks = seq(0, 1, by = 0.25),
  labels = scales::percent_format(accuracy = 1),
  expand = expansion(mult = c(0.01, 0.01))
)

rdBu_10 <- c(
  "#B2182B", "#D6604D", "#F4A582", "#FDDBC7", "#F7F7F7",
  "#D1E5F0", "#92C5DE", "#4393C3", "#2166AC", "#053061"
)

rdBu_10_temp <- c("#053061", "#2166AC","#4393C3", "#92C5DE", "#D1E5F0",
                  "#F7F7F7", "#FDDBC7", "#F4A582", "#D6604D", "#B2182B")

bin_levels <- c(
      "0–5", "5–10", "10–15", "15–20", "20–25",
      "25–30", "30–35", "35–40", "40–45", "45–50",
      "50–55", "55–60", "60–65", "65–70", "70–75",
      "75–80", "80–85", "85–90", "90–95", "95–100")
 
load_response_results <- function(response_var, region_name) {
  
  out_dir <- get_out_dir(response_var, region_name)
  
  contrasts_ensemble <- read_all_files(
  file.path(out_dir, "contrasts"),
  "^contrasts_run_.*\\.csv$"
  )
  
  year_effects_ensemble <- read_all_files(
    file.path(out_dir, "year_effects"),
    "^year_effects_run_.*\\.csv$"
  )
  
  spatial_component_ensemble <- read_all_files(
    file.path(out_dir, "spatial_component"),
    "^spatial_component_run_.*\\.csv$"
  )
  
  diagnostics_ensemble <- read_all_files(
    file.path(out_dir, "diagnostics"),
    "^diagnostics_run_.*\\.csv$"
  )
  
  metrics_ensemble <- read_all_files(
    file.path(out_dir, "metrics"),
    "^metrics_run_.*\\.csv$"
  )
  
  logs_ensemble <- read_all_files(
    file.path(out_dir, "logs"),
    "^log_run_.*\\.csv$"
  )
  
  fwrite(contrasts_ensemble, file.path(out_dir, "tables", "contrasts_ensemble_all_runs.csv"),  sep = ";",  dec = ",")
  fwrite(year_effects_ensemble, file.path(out_dir, "tables", "year_effects_ensemble_all_runs.csv"),  sep = ";",  dec = ",")
  fwrite(spatial_component_ensemble, file.path(out_dir, "tables", "spatial_component_ensemble_all_runs.csv"),  sep = ";",  dec = ",")
  fwrite(diagnostics_ensemble, file.path(out_dir, "tables", "diagnostics_ensemble_sampled_all_runs.csv"),  sep = ";",  dec = ",")
  fwrite(metrics_ensemble, file.path(out_dir, "tables", "metrics_ensemble_all_runs.csv"),  sep = ";",  dec = ",")
  fwrite(logs_ensemble, file.path(out_dir, "tables", "logs_ensemble_all_runs.csv"),  sep = ";",  dec = ",")
  
  contrasts_long <- contrasts_ensemble %>%
  pivot_longer(
    cols = starts_with("PF_minus_SF"),
    names_to = "stage",
    values_to = "difference"
  ) %>%
  mutate(
    stage = recode(
      stage,
      "PF_minus_SF_early" = "SF[e]",
      "PF_minus_SF_intermediate" = "SF[i]",
      "PF_minus_SF_advanced" = "SF[a]"
    ),
    stage = factor(stage, levels = c("SF[e]", "SF[i]", "SF[a]")),
    contrast_type = factor(
      contrast_type,
      levels = c("Spatially adjusted", "Current landscape")
    ),
    frac_bin = case_when(
      contrast_type == "Current landscape" ~ as.character(cut(
        frac_area,
        breaks = seq(0, 1, by = 0.05),
        labels = bin_levels,
        include.lowest = TRUE,
        right = TRUE
      )),
      TRUE ~ NA_character_
    ),
    frac_bin = factor(frac_bin, levels = bin_levels),
    frac_plot = case_when(
      contrast_type == "Spatially adjusted" ~ round(frac_area, 6),
      contrast_type == "Current landscape" ~ as.numeric(frac_bin)
    ),
    response = response_var
  )
  
  contrasts_summary <- contrasts_long %>%
    filter(!is.na(frac_plot), !is.na(difference)) %>%
    group_by(region, response, contrast_type, frac_plot, frac_bin, stage) %>%
    summarise(
      median = median(difference, na.rm = TRUE),
      lower = safe_quantile(difference, 0.025),
      upper = safe_quantile(difference, 0.975),
      n_runs = n_distinct(run),
      .groups = "drop"
    )
  
  fwrite(
  contrasts_summary,
  file.path(out_dir, "tables", "contrasts_summary_empirical_CI.csv"),  sep = ";",  dec = ",")
  
  year_summary <- year_effects_ensemble %>%
    mutate(response = response_var) %>%
    group_by(region, response, Ano) %>%
    summarise(
      median = median(year_effect, na.rm = TRUE),
      lower = safe_quantile(year_effect, 0.025),
      upper = safe_quantile(year_effect, 0.975),
      n_runs = sum(!is.na(year_effect)),
      .groups = "drop"
    )
  
  fwrite(
  year_summary,
  file.path(out_dir, "tables", "year_effects_summary_empirical_CI.csv"),  sep = ";",  dec = ",")
  
  spatial_summary <- spatial_component_ensemble %>%
    mutate(response = response_var) %>%
    group_by(region, response, Lon, Lat) %>%
    summarise(
      spatial_median = median(spatial_component, na.rm = TRUE),
      spatial_mean = mean(spatial_component, na.rm = TRUE),
      spatial_sd = sd(spatial_component, na.rm = TRUE),
      n_runs = n(),
      .groups = "drop"
    )
  
  fwrite(
  spatial_summary,
  file.path(out_dir, "tables", "spatial_component_summary.csv"),  sep = ";",  dec = ",")
  
  metrics_summary <- metrics_ensemble %>%
    group_by(region, response) %>%
    summarise(
      n_successful_runs = n(),
      
      CCC_median = median(CCC, na.rm = TRUE),
      CCC_lower = safe_quantile(CCC, 0.025),
      CCC_upper = safe_quantile(CCC, 0.975),
      
      RMSE_median = median(RMSE, na.rm = TRUE),
      RMSE_lower = safe_quantile(RMSE, 0.025),
      RMSE_upper = safe_quantile(RMSE, 0.975),
      
      MAE_median = median(MAE, na.rm = TRUE),
      MAE_lower = safe_quantile(MAE, 0.025),
      MAE_upper = safe_quantile(MAE, 0.975),
      
      Bias_median = median(Bias, na.rm = TRUE),
      Bias_lower = safe_quantile(Bias, 0.025),
      Bias_upper = safe_quantile(Bias, 0.975),
      
      R2_median = median(R2, na.rm = TRUE),
      R2_lower = safe_quantile(R2, 0.025),
      R2_upper = safe_quantile(R2, 0.975),
      
      .groups = "drop"
    )

  fwrite(
    metrics_summary,
    file.path(out_dir, "tables", "metrics_summary_empirical_CI.csv"),  sep = ";",  dec = ",")

  
  list(
    contrasts_summary = contrasts_summary,
    year_summary = year_summary,
    spatial_summary = spatial_summary
  )
}

```

#### Plots functions

```{r}

spatial_breaks <- list(
  Prec  = seq(-750, 750, by = 250),
  Evapo = seq(-150, 150, by = 50),
  Temp  = seq(-1.5, 1.5, by = 0.5)
)

spatial_labels <- list(
  Prec  = c("≤ -750", "-500", "-250", "0", "250", "500", "≥ 750"),
  Evapo = c("≤ -150", "-100", "-50", "0", "50", "100", "≥ 150"),
  Temp  = c("≤ -1.5", "-1", "-0.5", "0", "0.5", "1", "≥ 1.5")
)

spatial_cols <- list(
  Prec  = rdBu_10,
  Evapo = rdBu_10,
  Temp  = rdBu_10_temp
)

make_p_adj <- function(dat, response_var, show_legend = FALSE, first_col = FALSE) {
  
  unit <- response_units[[response_var]]
  label <- response_labels[[response_var]]
  
  ggplot(
    dat %>% filter(contrast_type == "Spatially adjusted"),
    aes(x = frac_plot, y = median, color = stage, fill = stage)
  ) +
    geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.35) +
    geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.18, color = NA) +
    geom_line(linewidth = 0.8) +
    x_forest_scale +
    scale_color_manual(
      values = cols_by_response[[response_var]],
      labels = parse(text = c("bold(SF[e])", "bold(SF[i])", "bold(SF[a])"))) +
    scale_fill_manual(
      values = cols_by_response[[response_var]],
      labels = parse(text = c("bold(SF[e])", "bold(SF[i])", "bold(SF[a])")),
      guide = "none") +
    labs(
      title = label,
      x = NULL,
      y = ifelse(
        first_col,
        paste0("Spatially adjusted\nPF − SF(", unit, ")"),
        paste0("\nPF − SF(", unit, ")")),
      color = NULL,
      fill = NULL
    ) +
    theme_bw(base_size = 10) +
    theme(
      text = element_text(face = "bold", family = "serif"),
      plot.title = element_text(hjust = 0.5, size = 12),
      axis.title = element_text(size = 10),
      axis.text = element_text(size = 9, color = "black"),
      legend.text = element_text(size = 9, face = "bold"),
      legend.key.width = unit(0.45, "cm"),
      legend.key.height = unit(0.25, "cm"),
      legend.margin = margin(t = -3, r = 0, b = -6, l = 0),
      legend.box.margin = margin(t = -6, r = 0, b = -8, l = 0),
      plot.margin = margin(t = 2, r = 6, b = 10, l = 4),
      legend.position = ifelse(show_legend, "top", "none"),
      legend.direction = "horizontal",
      panel.grid.minor = element_blank()
    )
}

make_p_land <- function(dat, response_var, show_legend = FALSE, first_col = FALSE) {
  
  unit <- response_units[[response_var]]
  
  ggplot(
    dat %>% filter(contrast_type == "Current landscape"),
    aes(x = frac_bin, y = median, color = stage, group = stage)
  ) +
    geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.35) +
    geom_line(linewidth = 0.8, linetype = "22") +
    scale_x_discrete(
      drop = FALSE,
      breaks = bin_levels[c(FALSE, TRUE)]) +
    scale_color_manual(
      values = cols_by_response[[response_var]],
      labels = parse(text = c("bold(SF[e])", "bold(SF[i])", "bold(SF[a])"))
    ) +
    labs(
      x = "Forest cover (%)",
      y = ifelse(
        first_col,
        paste0("Current landscape\nPF − SF(", unit, ")"),
        paste0("\nPF − SF(", unit, ")")
      ),
      color = NULL
    ) +
    theme_bw(base_size = 10) +
    theme(
      text = element_text(face = "bold", family = "serif"),
      axis.title = element_text(size = 10),
      axis.text.y = element_text(size = 9, color = "black"),
      axis.text.x = element_text(size = 9, color = "black", angle = 45, hjust = 1),
      legend.text = element_text(size = 9, face = "bold"),
      legend.key.width = unit(0.45, "cm"),
      legend.key.height = unit(0.25, "cm"),
      legend.margin = margin(t = -3, r = 0, b = -6, l = 0),
      legend.box.margin = margin(t = -6, r = 0, b = -8, l = 0),
      plot.margin = margin(t = 2, r = 6, b = 10, l = 4),
      legend.position = ifelse(show_legend, "top", "none"),
      legend.direction = "horizontal",
      panel.grid.minor = element_blank()
    )
}

make_p_year <- function(dat, response_var, first_col = FALSE) {
  
  unit <- response_units[[response_var]]
  
  ggplot(dat, aes(x = Ano, y = median)) +
    geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.35) +
    geom_ribbon(aes(ymin = lower, ymax = upper), fill = color_year[[response_var]], alpha = 0.18) +
    geom_line(linewidth = 0.85, color = color_year[[response_var]]) +
    geom_point(size = 1.5, color = color_year[[response_var]]) +
    scale_x_continuous(
      breaks = seq(2001, 2021, by = 5),
      limits = c(2001, 2021)
    ) +
    labs(
      x = NULL,
      y = ifelse(
        first_col,
        paste0("Year effect\n(", unit, ")"),
        paste0("\n(", unit, ")"))) +
    theme_bw(base_size = 10) +
    theme(
      text = element_text(face = "bold", family = "serif"),
      axis.title = element_text(size = 10),
      axis.text = element_text(size = 9, color = "black"),
      plot.margin = margin(t = 2, r = 4, b = 10, l = 4),
      panel.grid.minor = element_blank()
    )
}

make_p_spatial <- function(dat, response_var, first_col = FALSE) {
  
  unit <- response_units[[response_var]]
  
  ggplot(dat, aes(x = Lon, y = Lat, z = spatial_median)) +
    stat_summary_2d(fun = median, bins = 120) +
    coord_quickmap(expand = TRUE) +
    scale_fill_stepsn(
      colours = spatial_cols[[response_var]],
      breaks = spatial_breaks[[response_var]],
      labels = spatial_labels[[response_var]],
      limits = range(spatial_breaks[[response_var]]),
      oob = scales::squish,
      name = paste0("Spatial deviation (", unit, ")"),
      guide = guide_coloursteps(
        title.position = "top",
        title.hjust = 0.5,
        label.position = "bottom",
        barwidth = unit(5.0, "cm"),
        barheight = unit(0.32, "cm"))) +
    labs(x = "Longitude", 
         y = ifelse(first_col, "Latitude", "")) +
    theme_bw(base_size = 10) +
    theme(
      text = element_text(face = "bold", family = "serif"),
      axis.title = element_text(size = 10),
      axis.text = element_text(size = 9, color = "black"),
      legend.position = "top",
      legend.direction = "horizontal",
      legend.title = element_text(size = 9, face = "bold"),
      legend.text = element_text(size = 8.5),
      legend.box.margin = margin(t = -4, r = 0, b = -2, l = 0),
      legend.margin = margin(t = -4, r = 0, b = -4, l = 0),
      plot.margin = margin(t = 2, r = 2, b = 0, l = 2),
      panel.grid.minor = element_blank())
  
}

```

#### Results

```{r}

all_results <- lapply(response_vars, load_response_results, region_name = region_name)
names(all_results) <- response_vars


p_adj_list <- Map(
  function(v, fc)
    make_p_adj(
      all_results[[v]]$contrasts_summary,
      response_var = v,
      show_legend = TRUE,
      first_col = fc
    ),
  response_vars,
  c(TRUE, FALSE, FALSE)
)

p_land_list <- Map(
  function(v, fc)
    make_p_land(
      all_results[[v]]$contrasts_summary,
      response_var = v,
      show_legend = TRUE,
      first_col = fc
    ),
  response_vars,
  c(TRUE, FALSE, FALSE)
)

p_year_list <- Map(
  function(v, fc)
    make_p_year(
      all_results[[v]]$year_summary,
      response_var = v,
      first_col = fc
    ),
  response_vars,
  c(TRUE, FALSE, FALSE)
)

p_spatial_list <- Map(
  function(v, fc)
    make_p_spatial(
      all_results[[v]]$spatial_summary,
      response_var = v,
      first_col = fc
    ),
  response_vars,
  c(TRUE, FALSE, FALSE)
)

out_fig_dir <- file.path("fig")

p_adj_list
ggsave(
  file.path(out_fig_dir, paste0("figure_spat_adj_", region_name, ".png")),
  p_adj_list, width = 9, height = 5.5, dpi = 600, bg = "white")

p_land_list
ggsave(
  file.path(out_fig_dir, paste0("figure_curr_land_", region_name, ".png")),
  p_land_list, width = 9, height = 5.5, dpi = 600, bg = "white")

p_year_list
ggsave(
  file.path(out_fig_dir, paste0("figure_year_effect_", region_name, ".png")),
  p_year_list, width = 9, height = 5.5, dpi = 600, bg = "white")

p_spatial_list
ggsave(
  file.path(out_fig_dir, paste0("figure_spat_effect_", region_name, ".png")),
  p_spatial_list, width = 9, height = 5.5, dpi = 600, bg = "white")

```

#### Panel

```{r}

tag_plot <- function(p, tag) {
  p + labs(tag = tag) +
    theme(
      plot.tag = element_text(face = "bold", size = 13, family = "serif"),
      plot.tag.position = c(0, 1)
    )
}

p_adj_list <- Map(tag_plot, p_adj_list, c("A", "E", "I"))
p_land_list <- Map(tag_plot, p_land_list, c("B", "F", "J"))
p_year_list <- Map(tag_plot, p_year_list, c("C", "G", "K"))
p_spatial_list <- Map(tag_plot, p_spatial_list, c("D", "H", "L"))

fig_region <- 
  wrap_plots(p_adj_list, nrow = 1) /
  wrap_plots(p_land_list, nrow = 1) /
  wrap_plots(p_year_list, nrow = 1) /
  wrap_plots(p_spatial_list, nrow = 1,  widths = c(1, 1, 1)) +
  plot_layout(heights = c(1, 1, 0.9, 1.20)) +
  plot_annotation(
    title = region_name,
    theme = theme(
      plot.title = element_text(
        size = 16,
        face = "bold",
        family = "serif",
        hjust = 0.5
      )
    )
  ) &
  theme(
    text = element_text(face = "bold", family = "serif"),
    plot.tag = element_text(face = "bold", size = 13)
  )

fig_region

out_fig_dir <- file.path("fig")
dir.create(out_fig_dir, showWarnings = FALSE, recursive = TRUE)

ggsave(
  file.path(out_fig_dir, paste0("painels/figure_ensemble_", region_name, "_all_variables.png")),
  fig_region, width = 10, height = 12, dpi = 600, bg = "white")

ggsave(file.path(out_fig_dir, paste0("painels/figure_ensemble_", region_name, "_all_variables.pdf")), 
       plot = fig_region, width = 10, height = 12, bg = "white")

```

### Diagnostics 

#### Functions

```{r}

safe_quantile <- function(x, p) {
  if (all(is.na(x))) return(NA_real_)
  as.numeric(quantile(x, p, na.rm = TRUE))
}

read_all_files <- function(path, pattern) {
  files <- list.files(path, pattern = pattern, full.names = TRUE)
  if (length(files) == 0) return(data.table())
  rbindlist(lapply(files, fread), fill = TRUE)
}

get_out_dir <- function(response_var, region_name) {
  paste0("spamm_ensemble_", tolower(response_var), "_diagnostics_", region_name)
}


region_name <- "EBA"
response_vars <- c("Prec", "Evapo", "Temp")

response_labels <- c(
  Prec  = "Precipitation",
  Evapo = "Evapotranspiration",
  Temp  = "Land surface temperature"
)

response_units <- c(
  Prec  = "mm",
  Evapo = "mm",
  Temp  = "°C"
)

color_year <- c(
  Prec  = "midnightblue",
  Evapo = "#02401B",
  Temp  = "#7F3B08"
)


load_diag_results <- function(response_var, region_name) {
  
  out_dir <- get_out_dir(response_var, region_name)
  
  diagnostics_ensemble <- read_all_files(
    file.path(out_dir, "diagnostics"),
    "^diagnostics_run_.*\\.csv$"
  )
  
  metrics_ensemble <- read_all_files(
    file.path(out_dir, "metrics"),
    "^metrics_run_.*\\.csv$"
  )
  
  if (nrow(diagnostics_ensemble) == 0) {
    stop("No diagnostics files found for ", response_var, " in ", out_dir)
  }
  
  if (nrow(metrics_ensemble) == 0) {
    stop("No metrics files found for ", response_var, " in ", out_dir)
  }
  
  diagnostics_ensemble <- diagnostics_ensemble %>%
    mutate(
      response = response_var,
      variable_label = response_labels[[response_var]]
    )
  
  metrics_ensemble <- metrics_ensemble %>%
    mutate(
      response = response_var,
      variable_label = response_labels[[response_var]]
    )
  
  list(
    diagnostics = diagnostics_ensemble,
    metrics = metrics_ensemble
  )
}


theme_diag <- function() {
  theme_bw(base_size = 11) +
    theme(
      text = element_text(face = "bold", family = "serif"),
      plot.title = element_text(hjust = 0.5, size = 12),
      axis.title = element_text(size = 11),
      axis.text = element_text(size = 10, color = "black"),
      strip.text = element_blank(),
      strip.background = element_blank(),
      panel.grid.minor = element_blank(),
      aspect.ratio = 1,
      plot.margin = margin(t = 3, r = 5, b = 4, l = 5)
    )
}



sample_diag <- function(dat, n_max = 30000) {
  set.seed(123)
  as.data.table(dat)[
    ,
    .SD[sample(.N, min(.N, n_max))],
    by = response
  ]
}

```

#### Plots

```{r}

make_p_resid <- function(dat, response_var, first_col = FALSE) {
  
  unit <- response_units[[response_var]]
  label <- response_labels[[response_var]]
  col <- color_year[[response_var]]
  
  ggplot(dat, aes(x = fitted, y = residual)) +
    geom_point(alpha = 0.14, size = 0.22, color = col) +
    geom_smooth(method = "loess", se = FALSE, color = "black", linewidth = 0.55) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "grey35", linewidth = 0.35) +
    labs(
      title = label,
      x = paste0("Fitted (", unit, ")"),
      y = ifelse(first_col, paste0("Residuals\n(", unit, ")"), paste0("\n(", unit, ")"))
    ) +
    theme_diag()
}

make_p_qq <- function(dat, response_var, first_col = FALSE) {
  
  unit <- response_units[[response_var]]
  col <- color_year[[response_var]]
  
  ggplot(dat, aes(sample = residual)) +
    stat_qq(size = 0.25, alpha = 0.25, color = col) +
    stat_qq_line(linewidth = 0.55, color = "black") +
    labs(
      x = "Theoretical quantiles",
      y = ifelse( first_col,  paste0("Sample quantiles\n(", unit, ")"),  paste0("\n(", unit, ")"))) +
    theme_diag()
}

make_p_obs_pred <- function(dat, metrics, response_var, first_col = FALSE) {
  
  unit <- response_units[[response_var]]
  col <- color_year[[response_var]]
  
  ccc_med <- median(metrics$CCC, na.rm = TRUE)
  r2_med <- median(metrics$R2, na.rm = TRUE)
  
  label_txt <- paste0(
    "CCC = ", round(ccc_med, 2),
    "\nR² = ", round(r2_med, 2)
  )
  
  ggplot(dat, aes(x = observed, y = fitted)) +
    geom_point(alpha = 0.18, size = 0.25, color = col) +
    geom_smooth(method = "lm", se = FALSE, color = "black", linewidth = 0.55) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey35", linewidth = 0.35) +
    annotate(
      "text",
      x = -Inf, y = Inf,
      label = label_txt,
      hjust = -0.1, vjust = 1.5,
      size = 3.5,
      family = "serif",
      fontface = "bold"
    ) +
    labs(
      x = paste0("Observed (", unit, ")"),
      y = ifelse(first_col, paste0("Predicted\n(", unit, ")"), paste0("\n(", unit, ")"))
    ) +
    theme_diag()
}

make_p_accuracy <- function(metrics, response_var, first_col = FALSE) {
  
  col <- color_year[[response_var]]
  
  dat <- metrics %>%
    select(region, run, CCC, R2) %>%
    pivot_longer(
      cols = c(CCC, R2),
      names_to = "metric",
      values_to = "value"
    ) %>%
    mutate(
      metric = factor(metric, levels = c("CCC", "R2"), labels = c("CCC", "R²"))
    )
  
  ggplot(dat, aes(x = metric, y = value)) +
    geom_boxplot(
      width = 0.45,
      outlier.shape = NA,
      linewidth = 0.45,
      fill = "white",
      color = col
    ) +
    geom_jitter(
      width = 0.08,
      alpha = 0.65,
      size = 1.2,
      color = col
    ) +
    stat_summary(
      fun = median,
      geom = "point",
      shape = 23,
      size = 2.4,
      fill = "white",
      color = "black"
    ) +
    coord_cartesian(ylim = c(0, 1)) +
    scale_y_continuous(breaks = seq(0, 1, by = 0.25)) +
    labs(
      x = NULL,
      y = ifelse(first_col, "Agreement /\nexplained variance", "")
    ) +
    theme_diag()
}

make_p_error <- function(metrics, response_var, first_col = FALSE) {
  
  unit <- response_units[[response_var]]
  col <- color_year[[response_var]]
  
  dat <- metrics %>%
    select(region, run, RMSE, MAE) %>%
    pivot_longer(
      cols = c(RMSE, MAE),
      names_to = "metric",
      values_to = "value"
    ) %>%
    mutate(
      metric = factor(metric, levels = c("RMSE", "MAE"))
    )
  
  ggplot(dat, aes(x = metric, y = value)) +
    geom_boxplot(
      width = 0.45,
      outlier.shape = NA,
      linewidth = 0.45,
      fill = "white",
      color = col
    ) +
    geom_jitter(
      width = 0.08,
      alpha = 0.65,
      size = 1.2,
      color = col
    ) +
    stat_summary(
      fun = median,
      geom = "point",
      shape = 23,
      size = 2.4,
      fill = "white",
      color = "black"
    ) +
    labs(
      x = NULL,
      y = ifelse(first_col, paste0("Error metrics\n(", unit, ")"), paste0("\n(", unit, ")"))
    ) +
    theme_diag()
}

```

#### Panel

```{r}

diag_results <- lapply(response_vars, load_diag_results, region_name = region_name)
names(diag_results) <- response_vars

diag_sampled <- lapply(diag_results, function(x) {
  sample_diag(x$diagnostics, n_max = 30000)
})

metrics_list <- lapply(diag_results, function(x) {
  x$metrics
})


first_col_flags <- c(TRUE, FALSE, FALSE)

p_resid_list <- Map(
  function(v, fc) {
    make_p_resid(
      diag_sampled[[v]],
      response_var = v,
      first_col = fc
    )
  },
  response_vars,
  first_col_flags
)

p_qq_list <- Map(
  function(v, fc) {
    make_p_qq(
      diag_sampled[[v]],
      response_var = v,
      first_col = fc
    )
  },
  response_vars,
  first_col_flags
)

p_obs_list <- Map(
  function(v, fc) {
    make_p_obs_pred(
      diag_sampled[[v]],
      metrics = metrics_list[[v]],
      response_var = v,
      first_col = fc
    )
  },
  response_vars,
  first_col_flags
)

p_acc_list <- Map(
  function(v, fc) {
    make_p_accuracy(
      metrics_list[[v]],
      response_var = v,
      first_col = fc
    )
  },
  response_vars,
  first_col_flags
)

p_err_list <- Map(
  function(v, fc) {
    make_p_error(
      metrics_list[[v]],
      response_var = v,
      first_col = fc
    )
  },
  response_vars,
  first_col_flags
)


tag_plot <- function(p, tag) {
  p + labs(tag = tag) +
    theme(
      plot.tag = element_text(face = "bold", size = 13, family = "serif"),
      plot.tag.position = c(0, 1)
    )
}

p_resid_list <- Map(tag_plot, p_resid_list, c("A", "E", "I"))
p_qq_list    <- Map(tag_plot, p_qq_list,    c("B", "F", "J"))
p_obs_list   <- Map(tag_plot, p_obs_list,   c("C", "G", "K"))
p_err_list   <- Map(tag_plot, p_err_list,   c("D", "H", "L"))


fig_diag_region <-
  wrap_plots(
    c(
      p_resid_list,
      p_qq_list,
      p_obs_list,
      p_err_list
    ),
    ncol = 3,
    byrow = TRUE
  ) +
  plot_layout(
    widths = c(1, 1, 1),
    heights = c(1, 1, 1, 1)
  ) +
  plot_annotation(
    title = paste0(region_name),
    theme = theme(
      plot.title = element_text(
        size = 16,
        face = "bold",
        family = "serif",
        hjust = 0.5
      )
    )
  ) &
  theme(
    text = element_text(face = "bold", family = "serif"),
    plot.tag = element_text(face = "bold", size = 13)
  )

fig_diag_region

ggsave(
  file.path(out_fig_dir, paste0("painels/figure_diagnostics_", region_name, "_all_variables.png")),
  fig_diag_region, width = 10, height = 11.5, dpi = 600, bg = "white"
)

ggsave(file.path(out_fig_dir, paste0("painels/figure_diagnostics_", region_name, "_all_variables.pdf")), 
       plot = fig_diag_region, width = 10, height = 11.5, bg = "white")

```


## SBA

### Main figure 

#### Base functions

```{r}

safe_quantile <- function(x, p) {
  if (all(is.na(x))) return(NA_real_)
  as.numeric(quantile(x, p, na.rm = TRUE))
}

read_all_files <- function(path, pattern) {
  files <- list.files(path, pattern = pattern, full.names = TRUE)
  if (length(files) == 0) return(data.table())
  rbindlist(lapply(files, fread), fill = TRUE)
}

get_out_dir <- function(response_var, region_name) {
  paste0("spamm_ensemble_", tolower(response_var), "_diagnostics_", region_name)
}


region_name <- "SBA"
response_vars <- c("Prec", "Evapo", "Temp")

response_labels <- c(
  Prec = "Precipitation",
  Evapo   = "Evapotranspiration",
  Temp  = "Land surface temperature")

response_units <- c(
  Prec = "mm",
  Evapo   = "mm",
  Temp  = "°C")

cols_by_response <- list(
  Prec = c(
    "SF[e]" = "#87CEFA",
    "SF[i]" = "#1E90FF",
    "SF[a]" = "#0000FF"),
  Evapo = c(
    "SF[e]" = "lightgreen",
    "SF[i]" = "green1",
    "SF[a]" = "green4"),
  Temp = c(
    "SF[e]" = "#FEE0B6",
    "SF[i]" = "#FDB863",
    "SF[a]" = "#E08214"))

color_year <- c(
  Prec  = "midnightblue",
  Evapo = "#02401B",
  Temp  = "#7F3B08")

x_forest_scale <- scale_x_continuous(
  limits = c(0, 1),
  breaks = seq(0, 1, by = 0.25),
  labels = scales::percent_format(accuracy = 1),
  expand = expansion(mult = c(0.01, 0.01))
)

rdBu_10 <- c(
  "#B2182B", "#D6604D", "#F4A582", "#FDDBC7", "#F7F7F7",
  "#D1E5F0", "#92C5DE", "#4393C3", "#2166AC", "#053061"
)

rdBu_10_temp <- c("#053061", "#2166AC","#4393C3", "#92C5DE", "#D1E5F0",
                  "#F7F7F7", "#FDDBC7", "#F4A582", "#D6604D", "#B2182B")

bin_levels <- c(
      "0–5", "5–10", "10–15", "15–20", "20–25",
      "25–30", "30–35", "35–40", "40–45", "45–50",
      "50–55", "55–60", "60–65", "65–70", "70–75",
      "75–80", "80–85", "85–90", "90–95", "95–100")
 
load_response_results <- function(response_var, region_name) {
  
  out_dir <- get_out_dir(response_var, region_name)
  
  contrasts_ensemble <- read_all_files(
  file.path(out_dir, "contrasts"),
  "^contrasts_run_.*\\.csv$"
  )
  
  year_effects_ensemble <- read_all_files(
    file.path(out_dir, "year_effects"),
    "^year_effects_run_.*\\.csv$"
  )
  
  spatial_component_ensemble <- read_all_files(
    file.path(out_dir, "spatial_component"),
    "^spatial_component_run_.*\\.csv$"
  )
  
  diagnostics_ensemble <- read_all_files(
    file.path(out_dir, "diagnostics"),
    "^diagnostics_run_.*\\.csv$"
  )
  
  metrics_ensemble <- read_all_files(
    file.path(out_dir, "metrics"),
    "^metrics_run_.*\\.csv$"
  )
  
  logs_ensemble <- read_all_files(
    file.path(out_dir, "logs"),
    "^log_run_.*\\.csv$"
  )
  
  fwrite(contrasts_ensemble, file.path(out_dir, "tables", "contrasts_ensemble_all_runs.csv"),  sep = ";",  dec = ",")
  fwrite(year_effects_ensemble, file.path(out_dir, "tables", "year_effects_ensemble_all_runs.csv"),  sep = ";",  dec = ",")
  fwrite(spatial_component_ensemble, file.path(out_dir, "tables", "spatial_component_ensemble_all_runs.csv"),  sep = ";",  dec = ",")
  fwrite(diagnostics_ensemble, file.path(out_dir, "tables", "diagnostics_ensemble_sampled_all_runs.csv"),  sep = ";",  dec = ",")
  fwrite(metrics_ensemble, file.path(out_dir, "tables", "metrics_ensemble_all_runs.csv"),  sep = ";",  dec = ",")
  fwrite(logs_ensemble, file.path(out_dir, "tables", "logs_ensemble_all_runs.csv"),  sep = ";",  dec = ",")
  
  contrasts_long <- contrasts_ensemble %>%
  pivot_longer(
    cols = starts_with("PF_minus_SF"),
    names_to = "stage",
    values_to = "difference"
  ) %>%
  mutate(
    stage = recode(
      stage,
      "PF_minus_SF_early" = "SF[e]",
      "PF_minus_SF_intermediate" = "SF[i]",
      "PF_minus_SF_advanced" = "SF[a]"
    ),
    stage = factor(stage, levels = c("SF[e]", "SF[i]", "SF[a]")),
    contrast_type = factor(
      contrast_type,
      levels = c("Spatially adjusted", "Current landscape")
    ),
    frac_bin = case_when(
      contrast_type == "Current landscape" ~ as.character(cut(
        frac_area,
        breaks = seq(0, 1, by = 0.05),
        labels = bin_levels,
        include.lowest = TRUE,
        right = TRUE
      )),
      TRUE ~ NA_character_
    ),
    frac_bin = factor(frac_bin, levels = bin_levels),
    frac_plot = case_when(
      contrast_type == "Spatially adjusted" ~ round(frac_area, 6),
      contrast_type == "Current landscape" ~ as.numeric(frac_bin)
    ),
    response = response_var
  )
  
  contrasts_summary <- contrasts_long %>%
    filter(!is.na(frac_plot), !is.na(difference)) %>%
    group_by(region, response, contrast_type, frac_plot, frac_bin, stage) %>%
    summarise(
      median = median(difference, na.rm = TRUE),
      lower = safe_quantile(difference, 0.025),
      upper = safe_quantile(difference, 0.975),
      n_runs = n_distinct(run),
      .groups = "drop"
    )
  
  fwrite(
  contrasts_summary,
  file.path(out_dir, "tables", "contrasts_summary_empirical_CI.csv"),  sep = ";",  dec = ",")
  
  year_summary <- year_effects_ensemble %>%
    mutate(response = response_var) %>%
    group_by(region, response, Ano) %>%
    summarise(
      median = median(year_effect, na.rm = TRUE),
      lower = safe_quantile(year_effect, 0.025),
      upper = safe_quantile(year_effect, 0.975),
      n_runs = sum(!is.na(year_effect)),
      .groups = "drop"
    )
  
  fwrite(
  year_summary,
  file.path(out_dir, "tables", "year_effects_summary_empirical_CI.csv"),  sep = ";",  dec = ",")
  
  spatial_summary <- spatial_component_ensemble %>%
    mutate(response = response_var) %>%
    group_by(region, response, Lon, Lat) %>%
    summarise(
      spatial_median = median(spatial_component, na.rm = TRUE),
      spatial_mean = mean(spatial_component, na.rm = TRUE),
      spatial_sd = sd(spatial_component, na.rm = TRUE),
      n_runs = n(),
      .groups = "drop"
    )
  
  fwrite(
  spatial_summary,
  file.path(out_dir, "tables", "spatial_component_summary.csv"),  sep = ";",  dec = ",")
  
  metrics_summary <- metrics_ensemble %>%
    group_by(region, response) %>%
    summarise(
      n_successful_runs = n(),
      
      CCC_median = median(CCC, na.rm = TRUE),
      CCC_lower = safe_quantile(CCC, 0.025),
      CCC_upper = safe_quantile(CCC, 0.975),
      
      RMSE_median = median(RMSE, na.rm = TRUE),
      RMSE_lower = safe_quantile(RMSE, 0.025),
      RMSE_upper = safe_quantile(RMSE, 0.975),
      
      MAE_median = median(MAE, na.rm = TRUE),
      MAE_lower = safe_quantile(MAE, 0.025),
      MAE_upper = safe_quantile(MAE, 0.975),
      
      Bias_median = median(Bias, na.rm = TRUE),
      Bias_lower = safe_quantile(Bias, 0.025),
      Bias_upper = safe_quantile(Bias, 0.975),
      
      R2_median = median(R2, na.rm = TRUE),
      R2_lower = safe_quantile(R2, 0.025),
      R2_upper = safe_quantile(R2, 0.975),
      
      .groups = "drop"
    )

  fwrite(
    metrics_summary,
    file.path(out_dir, "tables", "metrics_summary_empirical_CI.csv"),  sep = ";",  dec = ",")

  
  list(
    contrasts_summary = contrasts_summary,
    year_summary = year_summary,
    spatial_summary = spatial_summary
  )
}

```

#### Plots functions

```{r}

spatial_breaks <- list(
  Prec  = seq(-750, 750, by = 250),
  Evapo = seq(-150, 150, by = 50),
  Temp  = seq(-1.5, 1.5, by = 0.5)
)

spatial_labels <- list(
  Prec  = c("≤ -750", "-500", "-250", "0", "250", "500", "≥ 750"),
  Evapo = c("≤ -150", "-100", "-50", "0", "50", "100", "≥ 150"),
  Temp  = c("≤ -1.5", "-1", "-0.5", "0", "0.5", "1", "≥ 1.5")
)

spatial_cols <- list(
  Prec  = rdBu_10,
  Evapo = rdBu_10,
  Temp  = rdBu_10_temp
)

make_p_adj <- function(dat, response_var, show_legend = FALSE, first_col = FALSE) {
  
  unit <- response_units[[response_var]]
  label <- response_labels[[response_var]]
  
  ggplot(
    dat %>% filter(contrast_type == "Spatially adjusted"),
    aes(x = frac_plot, y = median, color = stage, fill = stage)
  ) +
    geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.35) +
    geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.18, color = NA) +
    geom_line(linewidth = 0.8) +
    x_forest_scale +
    scale_color_manual(
      values = cols_by_response[[response_var]],
      labels = parse(text = c("bold(SF[e])", "bold(SF[i])", "bold(SF[a])"))) +
    scale_fill_manual(
      values = cols_by_response[[response_var]],
      labels = parse(text = c("bold(SF[e])", "bold(SF[i])", "bold(SF[a])")),
      guide = "none") +
    labs(
      title = label,
      x = NULL,
      y = ifelse(
        first_col,
        paste0("Spatially adjusted\nPF − SF(", unit, ")"),
        paste0("\nPF − SF(", unit, ")")),
      color = NULL,
      fill = NULL
    ) +
    theme_bw(base_size = 10) +
    theme(
      text = element_text(face = "bold", family = "serif"),
      plot.title = element_text(hjust = 0.5, size = 12),
      axis.title = element_text(size = 10),
      axis.text = element_text(size = 9, color = "black"),
      legend.text = element_text(size = 9, face = "bold"),
      legend.key.width = unit(0.45, "cm"),
      legend.key.height = unit(0.25, "cm"),
      legend.margin = margin(t = -3, r = 0, b = -6, l = 0),
      legend.box.margin = margin(t = -6, r = 0, b = -8, l = 0),
      plot.margin = margin(t = 2, r = 6, b = 10, l = 4),
      legend.position = ifelse(show_legend, "top", "none"),
      legend.direction = "horizontal",
      panel.grid.minor = element_blank()
    )
}

make_p_land <- function(dat, response_var, show_legend = FALSE, first_col = FALSE) {
  
  unit <- response_units[[response_var]]
  
  ggplot(
    dat %>% filter(contrast_type == "Current landscape"),
    aes(x = frac_bin, y = median, color = stage, group = stage)
  ) +
    geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.35) +
    geom_line(linewidth = 0.8, linetype = "22") +
    scale_x_discrete(
      drop = FALSE,
      breaks = bin_levels[c(FALSE, TRUE)]) +
    scale_color_manual(
      values = cols_by_response[[response_var]],
      labels = parse(text = c("bold(SF[e])", "bold(SF[i])", "bold(SF[a])"))
    ) +
    labs(
      x = "Forest cover (%)",
      y = ifelse(
        first_col,
        paste0("Current landscape\nPF − SF(", unit, ")"),
        paste0("\nPF − SF(", unit, ")")
      ),
      color = NULL
    ) +
    theme_bw(base_size = 10) +
    theme(
      text = element_text(face = "bold", family = "serif"),
      axis.title = element_text(size = 10),
      axis.text.y = element_text(size = 9, color = "black"),
      axis.text.x = element_text(size = 9, color = "black", angle = 45, hjust = 1),
      legend.text = element_text(size = 9, face = "bold"),
      legend.key.width = unit(0.45, "cm"),
      legend.key.height = unit(0.25, "cm"),
      legend.margin = margin(t = -3, r = 0, b = -6, l = 0),
      legend.box.margin = margin(t = -6, r = 0, b = -8, l = 0),
      plot.margin = margin(t = 2, r = 6, b = 10, l = 4),
      legend.position = ifelse(show_legend, "top", "none"),
      legend.direction = "horizontal",
      panel.grid.minor = element_blank()
    )
}

make_p_year <- function(dat, response_var, first_col = FALSE) {
  
  unit <- response_units[[response_var]]
  
  ggplot(dat, aes(x = Ano, y = median)) +
    geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.35) +
    geom_ribbon(aes(ymin = lower, ymax = upper), fill = color_year[[response_var]], alpha = 0.18) +
    geom_line(linewidth = 0.85, color = color_year[[response_var]]) +
    geom_point(size = 1.5, color = color_year[[response_var]]) +
    scale_x_continuous(
      breaks = seq(2001, 2021, by = 5),
      limits = c(2001, 2021)
    ) +
    labs(
      x = NULL,
      y = ifelse(
        first_col,
        paste0("Year effect\n(", unit, ")"),
        paste0("\n(", unit, ")"))) +
    theme_bw(base_size = 10) +
    theme(
      text = element_text(face = "bold", family = "serif"),
      axis.title = element_text(size = 10),
      axis.text = element_text(size = 9, color = "black"),
      plot.margin = margin(t = 2, r = 4, b = 10, l = 4),
      panel.grid.minor = element_blank()
    )
}

make_p_spatial <- function(dat, response_var, first_col = FALSE) {
  
  unit <- response_units[[response_var]]
  
  ggplot(dat, aes(x = Lon, y = Lat, z = spatial_median)) +
    stat_summary_2d(fun = median, bins = 120) +
    coord_quickmap(expand = TRUE) +
    scale_fill_stepsn(
      colours = spatial_cols[[response_var]],
      breaks = spatial_breaks[[response_var]],
      labels = spatial_labels[[response_var]],
      limits = range(spatial_breaks[[response_var]]),
      oob = scales::squish,
      name = paste0("Spatial deviation (", unit, ")"),
      guide = guide_coloursteps(
        title.position = "top",
        title.hjust = 0.5,
        label.position = "bottom",
        barwidth = unit(5.0, "cm"),
        barheight = unit(0.32, "cm"))) +
    labs(x = "Longitude", 
         y = ifelse(first_col, "Latitude", "")) +
    theme_bw(base_size = 10) +
    theme(
      text = element_text(face = "bold", family = "serif"),
      axis.title = element_text(size = 10),
      axis.text = element_text(size = 9, color = "black"),
      legend.position = c(0.5, 1.75),
      legend.justification = c(0.5, 1),
      legend.direction = "horizontal",
      legend.title = element_text(size = 9, face = "bold"),
      legend.text = element_text(size = 8.5),
      legend.background = element_blank(),
      legend.box.background = element_blank(),
      plot.margin = margin(t = 35, r = 2, b = 0, l = 2),
      panel.grid.minor = element_blank()
    )
  
}

```

#### Results

```{r}

all_results <- lapply(response_vars, load_response_results, region_name = region_name)
names(all_results) <- response_vars


p_adj_list <- Map(
  function(v, fc)
    make_p_adj(
      all_results[[v]]$contrasts_summary,
      response_var = v,
      show_legend = TRUE,
      first_col = fc
    ),
  response_vars,
  c(TRUE, FALSE, FALSE)
)

p_land_list <- Map(
  function(v, fc)
    make_p_land(
      all_results[[v]]$contrasts_summary,
      response_var = v,
      show_legend = TRUE,
      first_col = fc
    ),
  response_vars,
  c(TRUE, FALSE, FALSE)
)

p_year_list <- Map(
  function(v, fc)
    make_p_year(
      all_results[[v]]$year_summary,
      response_var = v,
      first_col = fc
    ),
  response_vars,
  c(TRUE, FALSE, FALSE)
)

p_spatial_list <- Map(
  function(v, fc)
    make_p_spatial(
      all_results[[v]]$spatial_summary,
      response_var = v,
      first_col = fc
    ),
  response_vars,
  c(TRUE, FALSE, FALSE)
)

out_fig_dir <- file.path("fig")

p_adj_list
ggsave(
  file.path(out_fig_dir, paste0("figure_spat_adj_", region_name, ".png")),
  p_adj_list, width = 9, height = 5.5, dpi = 600, bg = "white")

p_land_list
ggsave(
  file.path(out_fig_dir, paste0("figure_curr_land_", region_name, ".png")),
  p_land_list, width = 9, height = 5.5, dpi = 600, bg = "white")

p_year_list
ggsave(
  file.path(out_fig_dir, paste0("figure_year_effect_", region_name, ".png")),
  p_year_list, width = 9, height = 5.5, dpi = 600, bg = "white")

p_spatial_list
ggsave(
  file.path(out_fig_dir, paste0("figure_spat_effect_", region_name, ".png")),
  p_spatial_list, width = 9, height = 5.5, dpi = 600, bg = "white")

```

#### Panel

```{r}

tag_plot <- function(p, tag) {
  p + labs(tag = tag) +
    theme(
      plot.tag = element_text(face = "bold", size = 13, family = "serif"),
      plot.tag.position = c(0, 1)
    )
}

p_adj_list <- Map(tag_plot, p_adj_list, c("A", "E", "I"))
p_land_list <- Map(tag_plot, p_land_list, c("B", "F", "J"))
p_year_list <- Map(tag_plot, p_year_list, c("C", "G", "K"))
p_spatial_list <- Map(tag_plot, p_spatial_list, c("D", "H", "L"))

fig_region <- 
  wrap_plots(p_adj_list, nrow = 1) /
  wrap_plots(p_land_list, nrow = 1) /
  wrap_plots(p_year_list, nrow = 1) /
  wrap_plots(p_spatial_list, nrow = 1,  widths = c(1, 1, 1)) +
  plot_layout(heights = c(1, 1, 0.9, 1.20)) +
  plot_annotation(
    title = region_name,
    theme = theme(
      plot.title = element_text(
        size = 16,
        face = "bold",
        family = "serif",
        hjust = 0.5
      )
    )
  ) &
  theme(
    text = element_text(face = "bold", family = "serif"),
    plot.tag = element_text(face = "bold", size = 13)
  )

fig_region

out_fig_dir <- file.path("fig")
dir.create(out_fig_dir, showWarnings = FALSE, recursive = TRUE)

ggsave(
  file.path(out_fig_dir, paste0("painels/figure_ensemble_", region_name, "_all_variables.png")),
  fig_region, width = 10, height = 12, dpi = 600, bg = "white")

ggsave(file.path(out_fig_dir, paste0("painels/figure_ensemble_", region_name, "_all_variables.pdf")), 
       plot = fig_region, width = 10, height = 12, bg = "white")

```

### Diagnostics 

#### Functions

```{r}

safe_quantile <- function(x, p) {
  if (all(is.na(x))) return(NA_real_)
  as.numeric(quantile(x, p, na.rm = TRUE))
}

read_all_files <- function(path, pattern) {
  files <- list.files(path, pattern = pattern, full.names = TRUE)
  if (length(files) == 0) return(data.table())
  rbindlist(lapply(files, fread), fill = TRUE)
}

get_out_dir <- function(response_var, region_name) {
  paste0("spamm_ensemble_", tolower(response_var), "_diagnostics_", region_name)
}


region_name <- "SBA"
response_vars <- c("Prec", "Evapo", "Temp")

response_labels <- c(
  Prec  = "Precipitation",
  Evapo = "Evapotranspiration",
  Temp  = "Land surface temperature"
)

response_units <- c(
  Prec  = "mm",
  Evapo = "mm",
  Temp  = "°C"
)

color_year <- c(
  Prec  = "midnightblue",
  Evapo = "#02401B",
  Temp  = "#7F3B08"
)


load_diag_results <- function(response_var, region_name) {
  
  out_dir <- get_out_dir(response_var, region_name)
  
  diagnostics_ensemble <- read_all_files(
    file.path(out_dir, "diagnostics"),
    "^diagnostics_run_.*\\.csv$"
  )
  
  metrics_ensemble <- read_all_files(
    file.path(out_dir, "metrics"),
    "^metrics_run_.*\\.csv$"
  )
  
  if (nrow(diagnostics_ensemble) == 0) {
    stop("No diagnostics files found for ", response_var, " in ", out_dir)
  }
  
  if (nrow(metrics_ensemble) == 0) {
    stop("No metrics files found for ", response_var, " in ", out_dir)
  }
  
  diagnostics_ensemble <- diagnostics_ensemble %>%
    mutate(
      response = response_var,
      variable_label = response_labels[[response_var]]
    )
  
  metrics_ensemble <- metrics_ensemble %>%
    mutate(
      response = response_var,
      variable_label = response_labels[[response_var]]
    )
  
  list(
    diagnostics = diagnostics_ensemble,
    metrics = metrics_ensemble
  )
}


theme_diag <- function() {
  theme_bw(base_size = 11) +
    theme(
      text = element_text(face = "bold", family = "serif"),
      plot.title = element_text(hjust = 0.5, size = 12),
      axis.title = element_text(size = 11),
      axis.text = element_text(size = 10, color = "black"),
      strip.text = element_blank(),
      strip.background = element_blank(),
      panel.grid.minor = element_blank(),
      aspect.ratio = 1,
      plot.margin = margin(t = 3, r = 5, b = 4, l = 5)
    )
}



sample_diag <- function(dat, n_max = 30000) {
  set.seed(123)
  as.data.table(dat)[
    ,
    .SD[sample(.N, min(.N, n_max))],
    by = response
  ]
}

```

#### Plots

```{r}

make_p_resid <- function(dat, response_var, first_col = FALSE) {
  
  unit <- response_units[[response_var]]
  label <- response_labels[[response_var]]
  col <- color_year[[response_var]]
  
  ggplot(dat, aes(x = fitted, y = residual)) +
    geom_point(alpha = 0.14, size = 0.22, color = col) +
    geom_smooth(method = "loess", se = FALSE, color = "black", linewidth = 0.55) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "grey35", linewidth = 0.35) +
    labs(
      title = label,
      x = paste0("Fitted (", unit, ")"),
      y = ifelse(first_col, paste0("Residuals\n(", unit, ")"), paste0("\n(", unit, ")"))
    ) +
    theme_diag()
}

make_p_qq <- function(dat, response_var, first_col = FALSE) {
  
  unit <- response_units[[response_var]]
  col <- color_year[[response_var]]
  
  ggplot(dat, aes(sample = residual)) +
    stat_qq(size = 0.25, alpha = 0.25, color = col) +
    stat_qq_line(linewidth = 0.55, color = "black") +
    labs(
      x = "Theoretical quantiles",
      y = ifelse( first_col,  paste0("Sample quantiles\n(", unit, ")"),  paste0("\n(", unit, ")"))) +
    theme_diag()
}

make_p_obs_pred <- function(dat, metrics, response_var, first_col = FALSE) {
  
  unit <- response_units[[response_var]]
  col <- color_year[[response_var]]
  
  ccc_med <- median(metrics$CCC, na.rm = TRUE)
  r2_med <- median(metrics$R2, na.rm = TRUE)
  
  label_txt <- paste0(
    "CCC = ", round(ccc_med, 2),
    "\nR² = ", round(r2_med, 2)
  )
  
  ggplot(dat, aes(x = observed, y = fitted)) +
    geom_point(alpha = 0.18, size = 0.25, color = col) +
    geom_smooth(method = "lm", se = FALSE, color = "black", linewidth = 0.55) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey35", linewidth = 0.35) +
    annotate(
      "text",
      x = -Inf, y = Inf,
      label = label_txt,
      hjust = -0.1, vjust = 1.5,
      size = 3.5,
      family = "serif",
      fontface = "bold"
    ) +
    labs(
      x = paste0("Observed (", unit, ")"),
      y = ifelse(first_col, paste0("Predicted\n(", unit, ")"), paste0("\n(", unit, ")"))
    ) +
    theme_diag()
}

make_p_accuracy <- function(metrics, response_var, first_col = FALSE) {
  
  col <- color_year[[response_var]]
  
  dat <- metrics %>%
    select(region, run, CCC, R2) %>%
    pivot_longer(
      cols = c(CCC, R2),
      names_to = "metric",
      values_to = "value"
    ) %>%
    mutate(
      metric = factor(metric, levels = c("CCC", "R2"), labels = c("CCC", "R²"))
    )
  
  ggplot(dat, aes(x = metric, y = value)) +
    geom_boxplot(
      width = 0.45,
      outlier.shape = NA,
      linewidth = 0.45,
      fill = "white",
      color = col
    ) +
    geom_jitter(
      width = 0.08,
      alpha = 0.65,
      size = 1.2,
      color = col
    ) +
    stat_summary(
      fun = median,
      geom = "point",
      shape = 23,
      size = 2.4,
      fill = "white",
      color = "black"
    ) +
    coord_cartesian(ylim = c(0, 1)) +
    scale_y_continuous(breaks = seq(0, 1, by = 0.25)) +
    labs(
      x = NULL,
      y = ifelse(first_col, "Agreement /\nexplained variance", "")
    ) +
    theme_diag()
}

make_p_error <- function(metrics, response_var, first_col = FALSE) {
  
  unit <- response_units[[response_var]]
  col <- color_year[[response_var]]
  
  dat <- metrics %>%
    select(region, run, RMSE, MAE) %>%
    pivot_longer(
      cols = c(RMSE, MAE),
      names_to = "metric",
      values_to = "value"
    ) %>%
    mutate(
      metric = factor(metric, levels = c("RMSE", "MAE"))
    )
  
  ggplot(dat, aes(x = metric, y = value)) +
    geom_boxplot(
      width = 0.45,
      outlier.shape = NA,
      linewidth = 0.45,
      fill = "white",
      color = col
    ) +
    geom_jitter(
      width = 0.08,
      alpha = 0.65,
      size = 1.2,
      color = col
    ) +
    stat_summary(
      fun = median,
      geom = "point",
      shape = 23,
      size = 2.4,
      fill = "white",
      color = "black"
    ) +
    labs(
      x = NULL,
      y = ifelse(first_col, paste0("Error metrics\n(", unit, ")"), paste0("\n(", unit, ")"))
    ) +
    theme_diag()
}

```

#### Panel

```{r}

diag_results <- lapply(response_vars, load_diag_results, region_name = region_name)
names(diag_results) <- response_vars

diag_sampled <- lapply(diag_results, function(x) {
  sample_diag(x$diagnostics, n_max = 30000)
})

metrics_list <- lapply(diag_results, function(x) {
  x$metrics
})


first_col_flags <- c(TRUE, FALSE, FALSE)

p_resid_list <- Map(
  function(v, fc) {
    make_p_resid(
      diag_sampled[[v]],
      response_var = v,
      first_col = fc
    )
  },
  response_vars,
  first_col_flags
)

p_qq_list <- Map(
  function(v, fc) {
    make_p_qq(
      diag_sampled[[v]],
      response_var = v,
      first_col = fc
    )
  },
  response_vars,
  first_col_flags
)

p_obs_list <- Map(
  function(v, fc) {
    make_p_obs_pred(
      diag_sampled[[v]],
      metrics = metrics_list[[v]],
      response_var = v,
      first_col = fc
    )
  },
  response_vars,
  first_col_flags
)

p_acc_list <- Map(
  function(v, fc) {
    make_p_accuracy(
      metrics_list[[v]],
      response_var = v,
      first_col = fc
    )
  },
  response_vars,
  first_col_flags
)

p_err_list <- Map(
  function(v, fc) {
    make_p_error(
      metrics_list[[v]],
      response_var = v,
      first_col = fc
    )
  },
  response_vars,
  first_col_flags
)


tag_plot <- function(p, tag) {
  p + labs(tag = tag) +
    theme(
      plot.tag = element_text(face = "bold", size = 13, family = "serif"),
      plot.tag.position = c(0, 1)
    )
}

p_resid_list <- Map(tag_plot, p_resid_list, c("A", "E", "I"))
p_qq_list    <- Map(tag_plot, p_qq_list,    c("B", "F", "J"))
p_obs_list   <- Map(tag_plot, p_obs_list,   c("C", "G", "K"))
p_err_list   <- Map(tag_plot, p_err_list,   c("D", "H", "L"))


fig_diag_region <-
  wrap_plots(
    c(
      p_resid_list,
      p_qq_list,
      p_obs_list,
      p_err_list
    ),
    ncol = 3,
    byrow = TRUE
  ) +
  plot_layout(
    widths = c(1, 1, 1),
    heights = c(1, 1, 1, 1)
  ) +
  plot_annotation(
    title = paste0(region_name),
    theme = theme(
      plot.title = element_text(
        size = 16,
        face = "bold",
        family = "serif",
        hjust = 0.5
      )
    )
  ) &
  theme(
    text = element_text(face = "bold", family = "serif"),
    plot.tag = element_text(face = "bold", size = 13)
  )

fig_diag_region

ggsave(
  file.path(out_fig_dir, paste0("painels/figure_diagnostics_", region_name, "_all_variables.png")),
  fig_diag_region, width = 10, height = 11.5, dpi = 600, bg = "white")

ggsave(file.path(out_fig_dir, paste0("painels/figure_diagnostics_", region_name, "_all_variables.pdf")), 
       plot = fig_diag_region, width = 10, height = 11.5, bg = "white")

```

# End